---
title: "Gating Cells in cytoverse"
vignette: >
  %\VignetteIndexEntry{Gating Cells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Background

The purpose of flow cytometry is to make inferences regarding some cell type(s) of interest. Often this involves establishing/drawing a gating hierachy to sequentially filter down to the cell type(s) of interest. This process is very often done manually, and can be very labour intensive. Importantly, manual approach implies **large variation** when one person does it vs. the next, or even if the same person does it multiple times. The `cytoverse` offers a suite of tools to tackle this problem in a reproducible and programmatic manner.

## GatingSet and GatingHeirarchy

In the previous sections, we have seen (and worked with) `flowFrame`, `flowSet`, `cytoframe`, and `cytoset`. These objects hold the underlying data, allowing us to visualize it, manipulate it, etc.

In this section we work with `GatingSet` and `GatingHierarchy`. These objects, like the name suggests, store information regarding various gates and filters that we will generate. Importantly, we can save the `GatingSet` which will completely package the analysis as well as the **.fcs** files in an opensource format that can be shared, allowing reproducibility.

In **part 1** of this section, we will go over variety of native methods that exist in the `cytoverse` which analysts can utilize to gate their cells of interest.

In **part 2**, we will demonstrate how analysts can automate this process by utilizing a `gatingTemplate` (a csv file that can be used to define and build the hierarchy).

## Building the gating tree: manual and semi-supervised

Required libraries

```{r load_library, message=FALSE, echo = TRUE, eval=TRUE, message=FALSE}
library(flowWorkspace)
library(ggcyto)
# set ggcyto theme  
theme_set(theme_classic())
library(CytoverseBioc2023)
cache_workshop_data()
```

First we create a `GatingSet`.

```{r intro_gatingset, eval = T, echo=TRUE, message = FALSE}
# create a cytoset
fcs_files <- get_workshop_data("fcs_data/")
fcs_files_TNK <- dplyr::filter(fcs_files, grepl(pattern = "TNK-CR1",x = rname)) # simple matching
cs <- flowWorkspace::load_cytoset_from_fcs(files = fcs_files_TNK$rpath, # loading a set of FCS files as a cytoset
                                           pattern = "TNK-CR1")

# add mock metadata
set.seed(123)
meta <- data.frame(name = sampleNames(cs), 
                   status = "Healthy", 
                   panel = "T Cell",
                   mock_treatment = sample(x = c("Treated","Control"),size = 4,prob = c(0.5,0.5),replace = TRUE),
                   row.names = sampleNames(cs))
pData(cs) <- meta

# creating a GatingSet
gs <- flowWorkspace::GatingSet(cs)
gs
```

We have now created a `GatingSet` called **gs**. Also note that we have associated some (albeit mock) metadata to the `GatingSet`. We are able to leverage the metadata to help construct gates (if need be).

First, we revisit some of the concepts that we have seen previously: **compensation** and **transformation** as it pertains to `GatingSet`.

We can directly compensate the `GatingSet` object.

```{r comp_trans_gs, echo=TRUE, eval=TRUE, fig.height=10, fig.width=10, message=FALSE,warning=FALSE}
# compensate a GatingSet
spill <- keyword(gs[[1]],"$SPILLOVER") # extract spillover matrix stored within the file

gs <- compensate(gs,spill) # GatingSet will be compensated and stores the spill matrix as well

recompute(gs) # update the gs

# retrieve compensation information; return NULL is GatingSet was not compensated
gs_spill <- gs_get_compensations(gs[[1]])

# output is a rich spillover information
slot(gs_spill[[1]],"spillover")[1:4, 1:4]
```

As well we can transform the `GatingSet`.

For convenience, we are using a previously defined **transformerList** object. The **transformerList** object was extracted from the workspace file created by the authors of [this](https://doi.org/10.1038/s41467-022-34638-2) study and can be found [here](https://flowrepository.org/public_experiment_representations/5932).

```{r demo_transformation_gs, echo = TRUE, eval = TRUE, message=FALSE}
# transformation of GatingSet
t.rds <- dplyr::filter(get_workshop_data("/fj_wsp"), # a previously defined transformation
                       grepl(pattern = "transform", 
                             x = rname))$rpath
my_trans <- readRDS(file = t.rds)

# make into transformerList object
my_trans_list <- flowWorkspace::transformerList(from = names(my_trans),
                                                trans = my_trans)

gs <- flowWorkspace::transform(gs, my_trans_list) # transforms underlying data
```

## Adding gates and building a hierarchy {.tabset}

On this `GatingSet`, we will add various gates that identifies cells of interest. Each sample within a `GatingSet` is associated with a `GatingHierarchy` that stores information regarding various gates that we would have created and applied to the samples.

Below, we demonstrate various types of gates that could be defined **programmatically** or be estimated **automatically** that can be applied to the samples within a `GatingSet` to build a hierarchy.

*Note:* This process is informed by **visualization** of the data. As such, we will make ample use of the `ggcyto` library.

### Programmatic construction of gates

#### Rectangular gate

As the name suggests, we can explicitly define a `rectangular gate` by by defining a `matrix` where the columns are the ranges for markers in `x-axis` and `y-axis`. Only 1 dimension can also be provided. In this scenario, the constructed gate filter the events marginated over the dimension provided.

```{r example_manual_rect, echo = TRUE, eval=TRUE, message=FALSE}
## Example of rectangleGate
cd3_vis <- ggcyto(
  gs, subset = "root", aes(x = "CD3", y = "CD56")
)+
  geom_hex(bins = 256)+
  facet_wrap(~name)

# before
cd3_vis

# using rectangle gate to add T cell gate
cd3_rectangle <- matrix(c(140, 205, 0, 200),
                       nrow = 2,ncol = 2,
                       byrow = F,
                       dimnames = list(NULL, # rownames
                                       c("V510-A","U570-A"))) # colnames are channel names
cd3_rectangle_gate <- rectangleGate(.gate = cd3_rectangle,filterId = "CD3+ T cells")
cd3_vis+geom_gate(cd3_rectangle_gate)
```

When a single dimension is provided:

```{r example_manual_rect_2, echo = TRUE, eval=TRUE, message=FALSE}
# provide the range
cd3_range <- matrix(data = c(140,200))
# set the column name to the requirec channel
colnames(cd3_range) <- "V510-A"

# construct gate
cd3_range_gate <- rectangleGate(.gate = cd3_range,
                                filterId = "CD3+ Range") 

cd3_vis + geom_gate(cd3_range_gate)
```

#### Polygon gate

Similar to `Rectangular gate` we can define a `Polygon gate` by generating a `matrix` of **vertices**.

Here, each **row** of the matrix is the coordinate for a **vertex** and the columns are the channels.

```{r example_manual_poly, echo=TRUE,eval=TRUE,message=FALSE}
## Example of polygonGate
nkt_vis <- ggcyto(gs, subset = "root",
                  aes(x = "CD1d", y = "CD3"))+ # fuzzy matching of marker names
  geom_hex(bins = 256)+
  facet_wrap(~name)

# visualize
nkt_vis

# define coordinates
## coordinates are based on visualization!
nkt_poly <- matrix(c(
  115,140, # are arranged as x,y pair
  150,150,
  150,180,
  200,180,
  200,140), # coordinates
  ncol = 2,
  byrow = T, # indicates that add as x,y pair
  dimnames = list(NULL, # rownames 
                  c("R670-A","V510-A")) # colnames are channel names
)

# create a gate
nkt_poly_gate <- polygonGate(nkt_poly,
                             filterId = "NKT cells")
# visualize
nkt_vis + geom_gate(nkt_poly_gate)
```

Editing a gate after construction is straight forward.

```{r edit_poly_gate, echo=TRUE, eval=TRUE, message=FALSE}
# move up and scale
nkt_poly_gate_scale <- flowCore::transform_gate(nkt_poly_gate, # gate object
                                          dx = 1, # which dimension to shift and by how much
                                        # dy = 1,
                                        # scale = 2 # scales both dimensions equally
                                          scale = c(1.05,1.05) # individually scale each dimension
                                          )
nkt_vis+geom_gate(nkt_poly_gate_scale)
```

##### Exercise

1.  Construct a new `Rectangular gate` that can be used to filter **CD4- CD3+** population. *Hint*: Visualize the data and then define the best ranges.
2.  How would you rotate this gate? *Hint*: try running `help(flowCore::transform_gate)`.

### Automatic estimation of gates

The `gs_add_gating_method` call from `openCyto` library can be leveraged for automatic estimation of gates, as well as clean scripting! Available gating methods can be listed by calling `gt_list_methods()`.

Unlike with **programmatic** construction, `gs_add_gating_method` not only estimates, but also attaches the gate to the `GatingSet`.

#### Singlets gate

A `singlets` gate is mostly used for data clean up, by removing out `doublets`.

```{r singlet_gate_viz, echo = T, eval = T, message = FALSE }
# visualization
singlet_vis <- autoplot(gs_pop_get_data(gs,"root"), # gs_pop_get_data(gs, node) extracts the underlying data
                        x = "FSC-A",
                        y = "FSC-H",
                        bins = 256
                        )+
              facet_wrap(~sampleNames(gs)) # leverage the metadata that is saved within the GatingSet to facet plots 
singlet_vis
```

Now, we estimate and add!

```{r estimate_singlets_gate, echo = T, eval = T, message = FALSE}
# estimate and add
gs_add_gating_method(gs = gs, # gatingset
                     alias = "singlets", # name given to the population
                     pop = "+", # indicate whether events inside or outside the gate should be filtered
                     parent = "root", # where to attach this node
                     dims = "FSC-A,FSC-H", # dimensions used to estimate the gate
                     gating_method = "singletGate", # one of the available gating methods
                     gating_args = "wider_gate = FALSE", # arguments passed to singletGate
                     )
# visualize
singlet_vis + 
  geom_gate(gs_pop_get_gate(gs,"singlets"))+
  facet_wrap(~name)

```

#### Estimating cutpoint and ranges using `gate_quantile` and `gate_mindensity2`

`gate_quantile` and `gate_mindensity2` can be used to estimate a cut point in the data, allowing the gating of data to the **left(-)** or the **right(+)** of the cut point.

```{r additional_gate_examples, echo=TRUE, eval=TRUE, message=FALSE}
# calculate a live gate 
## Example of a quantile gate

## visualize
live_viz <- ggcyto(gs,
                   subset = "singlets",
                   aes(x = "Live", y = "FSC-A"))+
            geom_hex(bins = 128)+
            facet_wrap(~name)
live_viz

# add gate using gate_quantile
gs_add_gating_method(gs,
                     alias = "live",
                     pop = "-",
                     parent = "singlets",
                     dims = "U450-A",
                     gating_method = "gate_quantile",
                     gating_args = "probs = 0.95")

# add gate using gate_mindensity2
gs_add_gating_method(gs,
                     alias = "live_mindensity",
                     pop = "-",
                     parent = "singlets",
                     dims = "U450-A",
                     gating_method = "gate_mindensity2",
                     gating_args = "max = 100, gate_range = c(50,75)")

# visualize gates
live_viz+
  geom_gate(gs_pop_get_gate(gs, "live"), colour = "red")+
  geom_gate(gs_pop_get_gate(gs, "live_mindensity"), colour = "blue")

```

We can also use the cutpoints to estimate a rectangular gate!

```{r example_rect_gate, echo=TRUE,eval=TRUE,message=TRUE}
# conventional T cells
t_cell_vis <- ggcyto(
  gs, subset = "live",
  aes(x  = "TCR Va7_2", y = "CD161")
) + geom_hex(bins = 256)+
  facet_wrap(~name)

# visualize 
t_cell_vis

# estimate and add
gs_add_gating_method(gs,
                     alias = "MAIT Cells",
                     pop = "++",
                     parent = "live",
                     dims = "G660-A,V710-A",
                     gating_method = "gate_quantile",
                     gating_args = "probs = 0.95, min = 50, max = 200")
# visualize
t_cell_vis+
  geom_gate(gs_pop_get_gate(gs,"MAIT Cells"))
```

#### Estimating an Ellipsoid gate

`gate_flowclust_2d` can be used to identify and generate an `Ellipsoid gate` in a semi-supervised manner.

```{r example_flowclust,echo=TRUE, eval=TRUE,message=FALSE, warning = FALSE}
cd3_ellipse <- ggcyto(gs,
       subset = "live",
       aes(x = "CD4", y = "CD3"))+
  geom_hex(bins = 256)

# estimate and add
gs_add_gating_method(gs,
                     alias = "CD3+",
                     pop = "+",
                     parent = "live",
                     dims = "U785-A,V510-A",
                     gating_method = "gate_flowclust_2d",
                     gating_args = "K = 3, target = c(100,150),quantile = 0.9, plot = FALSE") #

# visualize
cd3_ellipse+
  geom_gate(gs_pop_get_gate(gs,"CD3+"))
```

##### Exercise

1.  How would you add a gate called `CD4+ T Cells` defined as: **CD4+** and **CD3+**?\
    *Hint:* Use the visualization above to provide an appropriate target.
2.  `gating_args` above indicates `K = 3, quantile = 0.9, plot = FALSE`. Try running the code block by indicating `plot = TRUE`. What is the outcome? How can you run this code block again? *Hint:* try running `help.search(pattern = "remove gating")`.


#### Quad gate

When there are population(s) of interest in all 4 quadrants, it is useful to estimate a `Quadg gate`. For this, we can once again utilize `gate_quantile` and `gate_mindensity2`.
*Note*: Pay attention to `alias` and `pop` arguments!

```{r viz_cd4_cd8,echo=TRUE, eval=TRUE,message=FALSE}
# plot subsets
t_subsets <- ggcyto(gs,subset = "CD3+",
                  aes(x = "CD45RA", y = "CCR7"))+
  geom_hex(bins = 256)+
  facet_wrap(~name)

# visualize
t_subsets

# estimate and add
gs_add_gating_method(gs,
                     alias = "*",
                     pop = "+/-+/-",
                     parent = "CD3+",
                     dims = "CD45RA,CCR7",
                     gating_method = "gate_quantile",
                     gating_args = "probs = 0.9, mins = 0")


# visualize (exercise)
# Fill with call to ggcyto
```

##### Exercise

1. How would you confirm that the attached gates have correct boundaries? 
*Hint*: Refer to the section on visualization.
2. `plot(gs)` is displaying many nodes, not all are useful. Try to delete `CCR7+` and re-do the visualization!

### Boolean Gates

We can use `booleanFilter` to simply **negate** a gated population! The resulting filter does not have a geometric representation.

```{r example_bool_filter, echo=TRUE, eval=TRUE, message=FALSE}
# add not MAIT gate
## Example of booleanFilter
not_mait <- booleanFilter(`!MAIT Cells`, filterId = "not_MAIT")

# add Boolean gate
gs_pop_add(gs, not_mait, parent = "live")

recompute(gs)

# visualize
t_cell_vis+
  geom_overlay(
    gs_pop_get_data(gs, y = "not_MAIT"),
    size = 0.3, 
    colour = "red"
  )
```

*Note:* We are also able to combine multiple gates to generate a `booleanFilter`.

Let's create by combining `not_MAIT` and `MAIT Cells`

```{r example_or_filter, echo=TRUE, eval=FALSE, message=FALSE}
# using an OR ('|') operator
all_cells <- booleanFilter(`not_MAIT | MAIT Cells`,
                               filterId = "All")
# add population
gs_pop_add(gs, all_cells, parent = "live")
recompute(gs)
```

#### Exercise

1. What does `gs_add_pop` do? What is the purpose of `recompute(gs)` above?
2. How would you create a `booleanFilter` that only filters out **naive** T cells defined as **CD45RA+CCR7+**? Can you visualize the result?

## Visualizing the Gating Hierachy

We have been creating and adding many gates to our `GatingSet`. A useful way to visualize all the gates and their relationship is by plotting the gating Hierarchy as a tree. This gives us an immediate summary of what nodes are present in our `GatingSet`.

```{r plot_gs_tree,echo=TRUE,eval=TRUE,message=FALSE}
# plot the gating tree
plot(gs, bool = TRUE)
```

Another useful approach is to visualize the gated data, with gates that we have generated. See [articles/5_ggcyto.html] for more.

```{r viz_gs, echo=TRUE, eval=TRUE,message=FALSE}
# visualize the full gating hierarchy
autoplot(gs[[1]],bins = 256, bool = TRUE)+
  ggcyto_par_set(limits = "data") # set data range to be determined by data
```

## Saving your work

Finally, we save your `GatingSet`. The `GatingSet` can be loaded back into R using `load_gs("path/to/a/folder")`.

```{r save_gs, echo=TRUE, eval=FALSE}
# save your work
save_gs(gs, path = "path/to/a/folder")
```

## Conclusion

In this section, we spent some time identifying various types of gates that are available in `cytoverse`. As well, we demonstrated how to programmatically create such gates, either manually (i.e. by defining ranges or vertices) or in a semi-automated and data driven manner.

In [**Part 2**](4_Gating_gatingTemplate.Rmd) we will take a look at how users can leverage a `gatingTemplate` generate gates as we have done here, but with minimal scripting. It is worth noting that while the use `gatingTemplate` minimizes scripting, it does not compromise on reproducibility!

Lastly, we also made ample use of the `ggcyto` library in order to generate visualizations that helped in QC'ing the data and the gates that we generated. We will go over `ggcyto` in more detail in [**Visualizations**](5_ggcyto.Rmd).
