---
title: "Gating Cells in cytoverse"
vignette: >
  %\VignetteIndexEntry{Gating Cells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Background

The purpose of flow cytometry is to make inferences regarding some cell type(s) of interest. Often this involves establishing/drawing a gating hierarchy to sequentially filter down to the cell type(s) of interest. This process is very often done manually, and can be very labour intensive. Importantly, manual approach implies **large variation** when one person does it vs. the next, or even if the same person does it multiple times. The `cytoverse` offers a suite of tools to tackle this problem in a reproducible and programmatic manner.

## GatingSet and GatingHeirarchy

In the previous sections, we have seen (and worked with) `cytoframe`, and `cytoset`. These objects hold the underlying data, allowing us to visualize it, manipulate it, etc.

In this section we work with `GatingSet` and `GatingHierarchy`. These objects, like the name suggests, store information regarding various gates and filters that we will generate. Importantly, we can save the `GatingSet` which will completely package the analysis as well as the **.fcs** files in an opensource format that can be shared, allowing reproducibility.

In **part 1** of this section, we will go over variety of native methods that exist in the `cytoverse` which analysts can utilize to gate their cells of interest.

In **part 2** of this sections, we will go over methods to extract gated data for cells/populations of interest.

In **part 3 (optional)**, we will demonstrate how analysts can automate this process by utilizing a `gatingTemplate` (a csv file that can be used to define and build the hierarchy).

## Creating a `GatingSet`

Required libraries

```{r load_library, message=FALSE, echo = TRUE, eval=TRUE, message=FALSE}
library(flowWorkspace)
library(ggcyto)
# set ggcyto theme  
theme_set(theme_classic())
library(CytoverseBioc2023)
cache_workshop_data()
```

To create a `GatingSet`, first load in a `cytoset`. Here, we are making use of the `cytoset` that we have created previously. 

```{r intro_gatingset, eval = T, echo=TRUE, message = FALSE}
# load cytoset
cs <- make_cytoset(only_TNK = TRUE)

# creating a GatingSet
gs <- flowWorkspace::GatingSet(cs)
gs
```

We have now created a `GatingSet` called **gs**.

### Exercise
1. How would you check the metadata associated with the `GatingSet`?
2. How would you subset `gs` to only include samples where **Treatment = Control**?

Now, we revisit some of the concepts that we have seen previously: **compensation** and **transformation** as it pertains to a `GatingSet`.

We can directly compensate the `GatingSet` object.

```{r comp_trans_gs, echo=TRUE, eval=TRUE, fig.height=10, fig.width=10, message=FALSE,warning=FALSE}
# compensate a GatingSet
spill <- keyword(gs[[1]],"$SPILLOVER") # extract spillover matrix stored within the file

gs <- compensate(gs,spill) # GatingSet will be compensated and stores the spill matrix as well

recompute(gs) # update the gs

# retrieve compensation information
gs_spill <- gs_get_compensations(gs[[1]])

# output is a rich spillover information
slot(gs_spill[[1]],"spillover")[1:4, 1:4]
```

As well we can transform the `GatingSet`.

For convenience, we are using a previously defined **transformerList** object. The **transformerList** object was extracted from the workspace file created by the authors of [this](https://doi.org/10.1038/s41467-022-34638-2) study and can be found [here](https://flowrepository.org/public_experiment_representations/5932).

<details>
```{r demo_transformation_gs, echo = TRUE, eval = TRUE, message=FALSE}
# transformation of GatingSet
t.rds <- dplyr::filter(
  get_workshop_data("/fj_wsp"), # a previously defined transformation
  grepl(
    pattern = "transform", 
    x = rname
  )
)$rpath
my_trans <- readRDS(file = t.rds)

# make into transformerList object
my_trans_list <- flowWorkspace::transformerList(
  from = names(my_trans),
  trans = my_trans
)

gs <- flowWorkspace::transform(gs, my_trans_list) # transforms underlying data
```
</details>

### Exercise

1. How would you be able to tell if a `GatingSet` has been compensated?
2. How would be able to tell if a `GatingSet` has been transformed?
*Hint*: Try `help.search("get_transformation")`.
3. Why might it be beneficial to compensate and transform a `GatingSet` rather than a `cytoframe` or a `cytoset`?

## A note about `GatingHierarchy`

You may have noticed above that some function calls have `gs_` and others have `gh_`.

`gs_` indicates that  `GatingSet` while `gh_` indicates a `GatingHierarchy`.

The `GatingHierarchy` is a data structure that stores the sample-wise gating information present in a `GatingSet`. In essence a `GatingSet` is a collection of `GatingHierarchy`. 

You can access the `GatingHierarchy` using the `[[` subset operation. 

### Exercise
1. Try running `gh_get_compensation(gs)`. Why do you get an error message? How would you fix this?
2. What is the difference between `gs[1]` and `gs[[1]]`?
3. Why are there multiple data structures to represent and hold flow cytometry data in the `cytoverse`?


## Adding gates and building a hierarchy {.tabset}

On this `GatingSet`, we will add various gates that identifies cells of interest. Each sample within a `GatingSet` is associated with a `GatingHierarchy` that stores information regarding various gates that we would have created and applied to the samples.

Below, we demonstrate various types of gates that could be estimated **automatically** or be defined **programmatically** which can be applied to the samples within a `GatingSet` to build a hierarchy.

*Note:* This process is informed by **visualization** of the data. As such, we will make ample use of the `ggcyto` library.

### Automatic estimation of gates

The `gs_add_gating_method` call from `openCyto` library can be leveraged for automatic estimation of gates, as well as clean scripting! Available gating methods can be listed by calling `gt_list_methods()`.

#### Singlets gate

A `singlets` gate is mostly used for data clean up, by removing out `doublets`.

```{r singlet_gate_viz, echo = T, eval = T, message = FALSE }
# visualization
singlet_vis <- autoplot(
  gs_pop_get_data(gs,"root"), # gs_pop_get_data(gs, node) extracts the underlying data
  x = "FSC-A",
  y = "FSC-H",
  bins = 256
  ) +
  facet_wrap(~sampleNames(gs)
  ) # leverage the metadata that is saved within the GatingSet to facet plots 
singlet_vis
```

Now, we estimate and add!

```{r estimate_singlets_gate, echo = T, eval = T, message = FALSE}
# estimate and add
gs_add_gating_method(
  gs = gs, # gatingset
  alias = "singlets", # name given to the population
  pop = "+", # indicate whether events inside or outside the gate should be filtered
  parent = "root", # where to attach this node
  dims = "FSC-A,FSC-H", # dimensions used to estimate the gate
  gating_method = "singletGate", # one of the available gating methods
  gating_args = "wider_gate = FALSE", # arguments passed to singletGate
)
# visualize
singlet_vis + 
  geom_gate(
    gs_pop_get_gate(
      gs,
      "singlets"
    )
  )+
  facet_wrap(~name)

```


#### Estimating cutpoint and ranges using `gate_quantile` and `gate_mindensity2`

`gate_quantile` and `gate_mindensity2` can be used to estimate a cut point in the data, allowing the gating of data to the **left(-)** or the **right(+)** of the cut point.

```{r additional_gate_examples, echo=TRUE, eval=TRUE, message=FALSE}
# calculate a live gate 
## Example of a quantile gate

## visualize
live_viz <- ggcyto(
  gs,
  subset = "singlets",
  aes(x = "Live", y = "FSC-A")
) +
geom_hex(bins = 128)+
facet_wrap(~name)
# live_viz

# add gate using gate_quantile
gs_add_gating_method(
  gs,
  alias = "live",
  pop = "-",
  parent = "singlets",
  dims = "U450-A",
  gating_method = "gate_quantile",
  gating_args = list(
    probs = 0.95
  )
)

# add gate using gate_mindensity2
gs_add_gating_method(
  gs,
  alias = "live_mindensity",
  pop = "-",
  parent = "singlets",
  dims = "U450-A",
  gating_method = "gate_mindensity2",
  gating_args = list(
    max = 100, 
    gate_range = c(50, 75)
  )
)

# visualize the 2 extimated gates
live_viz+
  geom_gate(gs_pop_get_gate(gs, "live"), colour = "red")+
  geom_gate(gs_pop_get_gate(gs, "live_mindensity"), colour = "blue")

```

We can also use the cutpoints to estimate a rectangular gate!

```{r example_rect_gate, echo=TRUE,eval=TRUE,message=TRUE}
# conventional T cells
t_cell_vis <- ggcyto(
  gs, subset = "live",
  aes(x  = "TCR Va7_2", y = "CD161")
) + geom_hex(bins = 256)+
  facet_wrap(~name)

# visualize 
# t_cell_vis

# estimate and add
gs_add_gating_method(
  gs,
  alias = "MAIT Cells",
  pop = "++",
  parent = "live",
  dims = "G660-A,V710-A",
  gating_method = "gate_quantile",
  gating_args = list(
    probs = 0.95, 
    min = 50, 
    max = 200
  )
)
# visualize
t_cell_vis+
  geom_gate(gs_pop_get_gate(gs,"MAIT Cells"))
```

#### Estimating an Ellipsoid gate

`gate_flowclust_2d` can be used to identify and generate an `Ellipsoid gate` in a semi-supervised manner.

```{r example_flowclust,echo=TRUE, eval=TRUE,message=FALSE, warning = FALSE}
cd3_ellipse <- ggcyto(gs,
       subset = "live",
       aes(x = "CD4", y = "CD3"))+
  geom_hex(bins = 256)

# estimate and add
gs_add_gating_method(
  gs,
  alias = "CD3+",
  pop = "+",
  parent = "live",
  dims = "U785-A,V510-A",
  gating_method = "gate_flowclust_2d",
  gating_args = list(
    K = 3,
    target = c(100,150),
    quantile = 0.9,
    plot = FALSE
  )
)

# visualize
cd3_ellipse+
  geom_gate(gs_pop_get_gate(gs,"CD3+"))
```

##### Exercise

1.  How would you add a gate called `CD4+ T Cells` defined as: **CD4+** and **CD3+**?\
    *Hint:* Use the visualization above to provide an appropriate target.
2.  `gating_args` above indicates `K = 3, quantile = 0.9, plot = FALSE`. Try running the code block by indicating `plot = TRUE`. What is the outcome? How can you run this code block again? *Hint:* try running `help.search(pattern = "remove gating")`.


#### Quad gate

When there are population(s) of interest in all 4 quadrants, it is useful to estimate a `quadrant gate` using `gate_quad_tmix`.
*Note*: Pay attention to `alias` and `pop` arguments!

```{r viz_cd4_cd8,echo=TRUE, eval=TRUE, message=FALSE}
# plot subsets
t_subsets <- ggcyto(
  gs,
  subset = "CD3+",
  aes(x = "CD45RA", y = "CCR7")
)+
geom_hex(bins = 256)+
facet_wrap(~name)

# visualize
# t_subsets

# estimate and add
gs_add_gating_method(
  gs,
  alias = "*",
  pop = "+/-+/-",
  parent = "CD3+",
  dims = "CD45RA,CCR7",
  gating_method = "gate_quad_sequential",
  gating_args = list(
    gFunc = "mindensity"
  ),
  collapseDataForGating = TRUE
)
```

##### Exercise

1. How would you confirm that the attached gates have correct boundaries? 
*Hint*: Refer to the section on visualization. Gates/Nodes are a combination of CD45RA[+/-]CCR7[+-] 
2. How would you remove the gates attached by `gs_add_gating_method`?
*Hint*: Try running `help(gs_add_gating_method)` and read the details.
<details>`gs_remove_gating_method` reverses the the results of `gs_add_gating_method` in a step wise manner.</details>


### Programmatic construction of gates

Suppose you would like to add a gate that has a particular shape or range that cannot be easily estimated using the automatic estimation approaches above. In this case, you could **programmatically** define a gate by either providing ranges, coordinates, or logical statements as demonstrated below.

#### Rectangular gate

As the name suggests, we can explicitly define a `rectangular gate` by by defining a `matrix` where the columns are the ranges for markers in `x-axis` and `y-axis`. Only 1 dimension can also be provided. In this scenario, the constructed gate filter the events marginated over the dimension provided.

```{r example_manual_rect, echo = TRUE, eval=TRUE, message=FALSE}
## Example of rectangleGate
cd3_vis <- ggcyto(
  gs, subset = "root", aes(x = "CD3", y = "CD56")
)+
  geom_hex(bins = 256)+
  facet_wrap(~name)

# before
# cd3_vis

# using rectangle gate to add T cell gate
cd3_rectangle <- matrix(
  c(140, 205, 0, 200),
  nrow = 2,
  ncol = 2,
  byrow = F,
  dimnames = list(
    c("min", "max"), # rownames
    c("V510-A","U570-A") # colnames are channel names
  )
) 
cd3_rectangle_gate <- rectangleGate(
  .gate = cd3_rectangle,
  filterId = "CD3+ T cells"
)
cd3_vis+geom_gate(cd3_rectangle_gate)
```

When a single dimension is provided:

```{r example_manual_rect_2, echo = TRUE, eval=TRUE, message=FALSE}
# provide the range
cd3_range <- matrix(data = c(140,200))
# set the column name to the required channel
colnames(cd3_range) <- "V510-A"

# construct gate
cd3_range_gate <- rectangleGate(
  .gate = cd3_range,
  filterId = "CD3+ Range"
) 

cd3_vis + geom_gate(cd3_range_gate)
```

To add this gate we call `gs_pop_add` like so:

```{r show_add_gate, echo = TRUE, eval=TRUE}
gs_pop_add(
  gs,
  parent = "live",
  gate = cd3_range_gate
)
```

Each call to `gs_pop_add` requires a `recompute(gs)` call. 

##### Exercise

1. Try to run the following code: 

<details>
```{r no_recompute, echo=TRUE, eval=FALSE}
no_recompute <- ggcyto(gs, 
       subset = "CD3+ Range",
       aes(x = "CD3", y = "CD4"))+
  geom_hex(bins = 256)
no_recompute
```
</details> 

Why do you get an error?

*Note*:<details>While `gs_add_pop` attaches the gates to the `GatingSet`, the index of the filtered events are not calculated until you call `recompute`.</details>

#### Polygon gate

Similar to `Rectangular gate` we can define a `Polygon gate` by generating a `matrix` of **vertices**.

Here, each **row** of the matrix is the coordinate for a **vertex** and the columns are the channels.

```{r example_manual_poly, echo=TRUE,eval=TRUE,message=FALSE}
## Example of polygonGate
nkt_vis <- ggcyto(gs, subset = "root",
                  aes(x = "CD1d", y = "CD3"))+ # fuzzy matching of marker names
  geom_hex(bins = 256)+
  facet_wrap(~name)

# visualize
# nkt_vis

# define coordinates
## coordinates are based on visualization!
nkt_poly <- matrix(
  c(
    115,140, # are arranged as x,y pair
    150,150,
    150,180,
    200,180,
    200,140 
  ), 
  ncol = 2,
  byrow = T, # indicates that add as x,y pair
  dimnames = list(
    NULL, # rownames have no meaning in polygonGates
    c("R670-A","V510-A") # colnames are channel names
  ) 
)

# create a gate
nkt_poly_gate <- polygonGate(
  nkt_poly,
  filterId = "NKT cells"
)
# visualize
nkt_vis + geom_gate(nkt_poly_gate)
```

Editing a gate after construction is straight forward.

```{r edit_poly_gate, echo=TRUE, eval=TRUE, message=FALSE}
# move up and scale
nkt_poly_gate_scale <- flowCore::transform_gate(
  nkt_poly_gate, # gate object
  dx = 1, # which dimension to shift and by how much
                                        # dy = 1,
                                        # scale = 2 # scales both dimensions equally
  scale = c(1.05,1.05) # individually scale each dimension
)
nkt_vis+geom_gate(nkt_poly_gate_scale)
```

##### Exercise

1.  Construct a new `Rectangular gate` that can be used to filter **CD4- CD3+** population. *Hint*: Visualize the data and then define the best ranges.
2.  How would you rotate this gate? *Hint*: try running `help(flowCore::transform_gate)`.


#### Boolean Gates

We can use `booleanFilter` to simply **negate** a gated population! The resulting filter does not have a geometric representation.

```{r example_bool_filter, echo=TRUE, eval=TRUE, message=FALSE}
# add not MAIT gate
## Example of booleanFilter
not_mait <- booleanFilter(`!MAIT Cells`, filterId = "not_MAIT")

# add Boolean gate
gs_pop_add(
  gs, 
  not_mait, 
  parent = "live"
)

recompute(gs)

# visualize
t_cell_vis+
  geom_overlay(
    gs_pop_get_data(gs, y = "not_MAIT"),
    size = 0.3, 
    colour = "red"
  )
```

*Note:* We are also able to combine multiple gates to generate a `booleanFilter`.

Let's create by combining `not_MAIT` and `MAIT Cells`

```{r example_or_filter, echo=TRUE, eval=FALSE, message=FALSE}
# using an OR ('|') operator
all_cells <- booleanFilter(
  `not_MAIT|MAIT Cells`,
  filterId = "All"
)
# add population
gs_pop_add(gs, all_cells, parent = "live")
recompute(gs)
```

#### Exercise

1. How would you create a `booleanFilter` that only filters out **naive** T cells defined as **CD45RA+CCR7+**? Can you visualize the result?
2. How would you remove an attached gate? Is `recompute` required after removal of a gate?
3. What happens if you remove a parent gate?

## Visualizing the Gating Hierachy

We have been creating and adding many gates to our `GatingSet`. A useful way to visualize all the gates and their relationship is by plotting the gating Hierarchy as a tree. This gives us an immediate summary of what nodes are present in our `GatingSet`.

```{r plot_gs_tree,echo=TRUE,eval=TRUE,message=FALSE}
# plot the gating tree
plot(gs, bool = TRUE)
```

Another useful approach is to visualize the gated data, with gates that we have generated. See the section of [visualization](6_ggcyto.html) for more.

```{r viz_gs, echo=TRUE, eval=TRUE,message=FALSE}
# visualize the full gating hierarchy
autoplot(
  gs[[1]],
  bins = 256, 
  bool = TRUE)+
  ggcyto_par_set(limits = "data") # set data range to be determined by data
```

## Saving your work

Finally, we save your `GatingSet`. The `GatingSet` can be loaded back into R using `load_gs("path/to/a/folder")`.

```{r save_gs, echo=TRUE, eval=FALSE}
# save your work
save_gs(gs, path = "path/to/a/folder")
```

## Conclusion

In this section, we spent some time identifying various types of gates that are available in `cytoverse`. As well, we demonstrated how to programmatically create such gates, either manually (i.e. by defining ranges or vertices) or in a semi-automated and data driven manner. 

Next, we go over methods to exctract the gated data. 

The [**(Optional) Part 3**](5_Gating_gatingTemplate.html) goes over how users can leverage a `gatingTemplate` to generate gates as we have done here, but with minimal scripting. It is worth noting that while the use `gatingTemplate` minimizes scripting, it does not compromise on reproducibility!

Lastly, we also made ample use of the `ggcyto` library in order to generate visualizations that helped in QC'ing the data and the gates that we generated. We will go over `ggcyto` in more detail in [**Visualizations**](6_ggcyto.html).
