---
title: "Reporting: Counts, Frequency, etc"
vignette: >
  %\VignetteIndexEntry{Reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Background

The final step in the analysis of flow cytometry data is **Reporting**. Below, we show how we can quickly extract counts, frequencies, and expression values for their population(s) of interest.

```{r libraries, echo=T, eval=T, message=FALSE, warning=FALSE}
library(flowCore)
library(flowWorkspace)
library(ggcyto)
library(CytoverseBioc2023)
theme_set(
  theme_bw()
)
cache_workshop_data()
```

```{r get_gatingset, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
gs <- make_transformed_gs(add_gates = TRUE)
```

## Counts, Frequencies, and MFI

We can easily extract Counts and Frequencies of the gated cell population from the `GatingSet`. 

```{r counts, echo = TRUE, eval = TRUE, message = FALSE}
# get count for all
counts <- gs_pop_get_count_fast(gs,  # indicate which GatingSet to use
                                statistic = "count", # indicate what statistic is requested
                                format = "wide") # indicate if data should be in wide (samples are columns) or long (each node for each sample gets a row) format

knitr::kable(head(counts,20), caption = "Event counts")
```

Changing `statistic` to `freq` return frequency of the gated population with respect to the parent population. 

```{r frequency, echo = TRUE, eval = TRUE, message = FALSE}
freq <- gs_pop_get_count_fast(gs,
                              statistic = "freq",
                              format = "wide")
knitr::kable(head(freq,20), caption = "Event frequencies relative to parent node")
```

What if one wants to report frequencies relative to a different population?

```{r plot_gating_hierarchy, echo=TRUE, eval=TRUE, message = FALSE}
plot(gs, bool = TRUE)
```

Let's consider the scenario here: you would like to calculate the frequency of **MAIT Cells**, and **True_NKT Cells** relative to total **CD3+ Cells**.

We can easily achieve this by using `gs_pop_get_stats` and indicating which node(s) we want

```{r grand_parent_populations, echo=TRUE, eval=TRUE, message=FALSE}
library(magrittr)
# get count for specified nodes and make wider 
pop_counts <- gs_pop_get_stats(gs,
                                    node = c("CD3+ T cells","MAIT Cells", "NKT cells"),
                                    type = "count"
                                    ) %>% # output is a long column
  tidyr::pivot_wider(names_from = pop, 
                     values_from = count,names_prefix = "count_") %>% # convert to wide
  dplyr::mutate(MAIT_freq = `count_MAIT Cells`/`count_CD3+ T cells`,
         NKT_freq = `count_NKT cells`/`count_CD3+ T cells`)


knitr::kable(pop_counts,caption = "Frequency of MAIT and NKT Cells (relative to CD3+ T Cells)")
```

Another common statistic that is reported if MFI (median fluorescence intensity). To extract MFI, we can again make use of `gs_pop_get_stat`. It is important to realize that MFI is often reported in *raw/linear scale*. 

```{r get_MFI, eval=TRUE, echo=TRUE, message=FALSE}
mfi_dataframe <- gs_pop_get_stats(gs,
                 nodes = c("CD3+ T cells", "MAIT Cells", "CD4+CD8a-", "CD4-CD8a+"),
                 type = pop.MFI,
                 inverse.transform = TRUE)

mfi_dataframe[1:5, 1:5]
```

For convenience, MFI are extracted for all markers for each of the specified population. 

## A simple report

In the table above, we noticed that expression of **PD1** is higher in MAIT Cells. Below, we create 2 plots to help drive this point (if this is of interest). We will make use of `ggcyto` and `ggplot2`

### Boxplot

```{r dotplot_pd1, eval=TRUE, echo=TRUE, message=FALSE}
# plot some data
pd1_mfi_plot <- dplyr::left_join(pData(gs),mfi_dataframe, 
          by = c("name" = "sample")) %>% 
  ggplot(aes(x = mock_treatment, y = `PD1 BB660`, fill = pop))+
  geom_boxplot()+
  labs(x = "Cell Type", y = "median PD1 expression")+
  scale_fill_manual(values = c("CD3+ T cells" = "black",
                               "MAIT Cells" = "red",
                               "CD4+CD8a-" = "blue",
                               "CD4-CD8a+" = "green"),
                    )+
  theme(aspect.ratio = 1, 
        legend.position = "right")

pd1_mfi_plot
```

### Overlaid histogram

```{r histogram_overlay, eval = TRUE, echo = TRUE}
pd1_overlay <- ggcyto(gs,
       subset = "MAIT Cells",
       aes(x = "PD1"))+
  geom_density(fill = "red", alpha = 0.3)+ # MAIT Cells in red
  geom_overlay(
    data = gs_pop_get_data(gs,"CD3+ T cells"),
    fill = "black", alpha = 0.5)+ # CD3+ Cells in black
   geom_overlay(
    data = gs_pop_get_data(gs,"CD4+CD8a-"),
    fill = "blue", alpha = 0.5)+ # CD4+ T Cells in blue
   geom_overlay(
    data = gs_pop_get_data(gs,"CD4-CD8a+"),
    fill = "green", alpha = 0.5)+ # CD8+ T Cells in green
  axis_x_inverse_trans()+
  facet_wrap(mock_treatment~name, nrow = 1)+
  labs(title = "",x = "PD1")+
  theme(aspect.ratio = 1)
pd1_overlay <- as.ggplot(pd1_overlay)
```

### Final Figure

```{r final_figure, eval = TRUE, echo=TRUE, message=FALSE}
gridExtra::grid.arrange(
 pd1_mfi_plot,
 pd1_overlay,
 nrow = 2,
 left = "PD1 expression in T cell subsets"
)

```

It is now clear that the expression of PD1 indeed tends to be higher on MAIT Cells compared to non-MAIT T Cells. 

## (Optional) Additional statistics

It is also possible to calculate additional statistics for the population of interest. For instance, if you were interested in nth percentile expression of a specific marker you could define a function and use it in `gs_pop_get_stats` like so:

```{r user_stats_function, eval=TRUE, echo=TRUE, message=TRUE}
# define a function
my_quantile <- function(fr,percentile,chnl){
  matched_chnl <- flowCore::getChannelMarker(fr,chnl) # match channel name
  res <- apply(exprs(fr)[,matched_chnl[["name"]],drop = FALSE], 2, quantile,percentile) # get quantile for specific channel
  names(res) <- matched_chnl[["desc"]]
  return(res)
}

# get stats
median_cd4 <- gs_pop_get_stats(gs,
                 c("lymphocytes","CD3+ T cells","CD4+CD8a-","CD4-CD8a+"), 
                 type = my_quantile,
                 inverse.transform = TRUE,
                 stats.fun.arg = list(percentile = .5,
                                      chnl = "cd4")) 
knitr::kable(median_cd4,caption = "Median expression of CD4 extracted using user defined function")

```

As we see, we were able to extract median expression of the marker **CD4** on 4 different populations. As expected, expression is highest on **CD4+ T Cells** and lowest on **CD8+ T Cells**.

# Conclusion

In this section, we show how to easily generate reports and visualizations from the `GatingSet` object which we created earlier. We demonstrated some approaches to easily extract counts, frequencies, MFI, or a user defined metric from the `GatingSet` object easily. 
Also, as noted in the section [Visualizations](articles/ggcyto_1-5.html), the `ggcyto` makes use of the `ggplot2` framework providing a familiar user interface. Importantly, this allows users to prepare intricate figures for events within the `GatingSet`.
