---
title: "Gating Cells in cytoverse: Part 1"
vignette: >
  %\VignetteIndexEntry{Gating Cells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Background

The purpose of flow cytometry is to make inferences regarding some cell type(s) of interest. Often this involves establishing/drawing a gating hierachy to sequentially filter down to the cell type(s) of interest. This process is very often done manually, and can be very labour intensive. Importantly, manual approach implies **large variation** when one person does it vs. the next, or even if the same person does it multiple times. The `cytoverse` offers a suite of tools to tackle this problem in a reproducible and programmatic manner.

## GatingSet and GatingHeirarchy

In the previous sections, we have seen (and worked with) `flowFrame`, `flowSet`, `cytoframe`, and `cytoset`. These objects hold the underlying data, allowing us to visualize it, manipulate it, etc.

In this section we work with `GatingSet` and `GatingHierarchy`. These objects, like the name suggests, store information regarding various gates and filters that we will generate. Importantly, we can save the `GatingSet` which will completely package the analysis as well as the **.fcs** files in an opensource format that can be shared, allowing reproducibility.

In **part 1** of this section, we will go over variety of native methods that exist in the `cytoverse` which analysts can utilize to gate their cells of interest.

In **part 2**, we will demonstrate how analysts can automate this process by utilizing a `gatingTemplate` (a csv file that can be used to define and build the hierarchy).

## Building the gating tree: manual and semi-supervised

Required libraries

```{r load_library, message=FALSE, echo = TRUE, eval=TRUE, message=FALSE}
library(flowWorkspace)
library(ggcyto)
# set ggcyto theme  
theme_set(theme_classic())
library(CytoverseBioc2023)
cache_workshop_data()
```

First we create a `GatingSet`.

```{r intro_gatingset, eval = T, echo=TRUE, message = FALSE}
# create a cytoset
fcs_files <- get_workshop_data("fcs_data/")
fcs_files_TNK <- dplyr::filter(fcs_files, grepl(pattern = "TNK-CR1",x = rname))
cs <- flowWorkspace::load_cytoset_from_fcs(files = fcs_files_TNK$rpath, # loading a set of FCS files as a cytoset
                                           pattern = "TNK-CR1")

# add mock metadata
set.seed(123)
meta <- data.frame(name = sampleNames(cs), 
                   status = "Healthy", 
                   panel = "T Cell",
                   mock_treatment = sample(x = c("Treated","Control"),size = 4,prob = c(0.5,0.5),replace = TRUE),
                   row.names = sampleNames(cs))
pData(cs) <- meta

# creating a GatingSet
gs <- flowWorkspace::GatingSet(cs)
gs
```

We have now created a `GatingSet` called **gs**. Also note that we have associated some (albeit mock) metadata to the `GatingSet`. We are able to leverage the metadata to help construct gates (if need be).

First, we revisit some of the concepts that we have seen previously: **compensation** and **transformation**.

We can directly compensate the `GatingSet` object.

```{r comp_trans_gs, echo=TRUE, eval=TRUE, fig.height=10, fig.width=10, message=FALSE,warning=FALSE}
# compensate a GatingSet
spill <- keyword(gs[[1]],"$SPILLOVER") # extract spillover matrix stored within the file

gs <- compensate(gs,spill) # GatingSet will be compensated and stores the spill matrix as well

recompute(gs) # update the gs

# retrieve compensation information; return NULL is GatingSet was not compensated
gs_spill <- gs_get_compensations(gs[[1]])

# output is a rich spillover information
slot(gs_spill[[1]],"spillover")[1:4, 1:4]
```

As well we can transform the `GatingSet`.

For convenience, we are using a previously defined **transformerList** object. The **transformerList** object was extracted from the workspace file created by the authors of [this](https://doi.org/10.1038/s41467-022-34638-2) study and can be found [here](https://flowrepository.org/public_experiment_representations/5932).

```{r demo_transformation_gs, echo = TRUE, eval = TRUE, message=FALSE}
# transformation of GatingSet
t.rds <- dplyr::filter(get_workshop_data("/fj_wsp"), # a previously defined transformation
                       grepl(pattern = "transform", 
                             x = rname))$rpath
my_trans <- readRDS(file = t.rds)

# make into transformerList object
my_trans_list <- flowWorkspace::transformerList(from = names(my_trans),
                                                trans = my_trans)

gs <- flowWorkspace::transform(gs, my_trans_list) # transforms underlying data
```

## Adding gates and building a hierarchy

Now that we have created a `GatingSet` we should start adding various gates that identifies cells of interest. Each sample within a `GatingSet` is associated with a `GatingHierarchy` that stores information regarding various gates that we would have created and applied to the samples.

Below, we demonstrate various types of gates that could be estimated **automatically** or defined **manually** that can be applied to the samples within a `GatingSet` to build a hierarchy.

*Note:* This process is informed by **visualization** of the data. As such, we will make ample use of the `ggcyto` library.

### Singlets gate

Singlets gate is mostly used for data clean up.

```{r singlet_gate_viz, echo = T, eval = T, message = FALSE }
# visualization
singlet_vis <- autoplot(gs_pop_get_data(gs,"root"), # gs_pop_get_data(gs, node) extracts the underlying data
                        x = "FSC-A",
                        y = "FSC-H",
                        bins = 256
                        )+
              facet_wrap(~sampleNames(gs)) # leverage the metadata that is saved within the GatingSet to facet plots 

singlet_vis
```

Now, we estimate a gate that will be used to filter out **singlets**. For this, we make use of the `flowStats` library from the `cytoverse`.

```{r estimate_singlets_gate, echo = T, eval = T, message = FALSE}
# estimate
singlet_gate <- fsApply(gs_pop_get_data(gs),function(x){ # sample wise calculation using fsApply
                flowStats::gate_singlet(x, 
                                        filterId = "singlet", # name for the gate
                                        wider_gate = FALSE, # indicate if the returned gate should be linient
                                        prediction_level = 0.95,
                                        subsample_pct = 0.5 # percent of events to sample to calculate the gate
                                        )
  })

# visualize
singlet_vis + 
  geom_gate(singlet_gate)+ # geom_gate takes a single gate or a list of gates as an argument. list should have the same name as sampleNames(gs)
  facet_wrap(~name)

```

Once visualized, we can add the gates to its respective samples. *Note:* Singlets gate was estimated for each sample.

```{r add_singlets_gate_2, eval=T, echo=TRUE, message = FALSE}
# add gate to the GatingSet 
gs_pop_add(gs, singlet_gate, parent = "root") # add singlet_gate to the root node
recompute(gs) # recompute updates the gs
```

### Estimating cutpoint using `gate_quantile`

`gate_quantile` can be used to estimate a cut point at a given quantile allowing the gating of data to the left or the right of the cut point.

```{r additional_gate_examples, echo=TRUE, eval=TRUE, message=FALSE}
# calculate a live gate 
## Example of a quantile gate

## visualize
live_viz <- ggcyto(gs,
                   subset = "singlet",
                   aes(x = "Live"))+
            geom_histogram(bins = 128)+
            facet_wrap(~name)
live_viz
  
live_gate <- fsApply(gs_pop_get_data(gs,"singlet"), # indicating that estimation should be done on filtered data: singlet
                     function(x){
                       gate <- openCyto::gate_quantile(fr = x, channel = "U450-A", 
                                                      probs = 0.95,
                                                      filterId = "live"
                                                      )
                       
                       # keep negative events
                       gate@max <- gate@min
                       gate@min <- -Inf
                       return(gate)
                     })

# visualize gate
live_viz+
  geom_gate(live_gate)

# add live gate
gs_pop_add(gs, parent = "singlet",gate = live_gate)
recompute(gs)
```

### Ellipsoid gate

Here, we are using `gate_flowclust_2d` from the `openCyto` library to estimate the location of **lymphocytes**. The output is an `Ellipsoid gate`.

```{r flowclust_2d_lymphocyte, echo=TRUE, eval=T, message=FALSE}
# calculate lymphocyte gate
## Example of using flowClust

## visualize
lymph_viz <- ggcyto(gs,
                   subset = "live",
                   aes(x = "FSC-A",
                       y = "SSC-A"))+
            geom_hex(bins = 128)

lymph_viz

# define
lymphocyte_gate <- fsApply(gs_pop_get_data(gs,"live"),
                           function(x){
                             openCyto::gate_flowclust_2d(fr = x,
                                                         xChannel = "FSC-A",
                                                         yChannel = "SSC-A",
                                                         filterId = "lymphocytes",
                                                         K = 1,target = c(1E5,0.5E3))
                           })


# call it lymphocytes
lymphocyte_gate <- lapply(lymphocyte_gate,
                          function(x){
                            x@filterId <- "lymphocytes"
                            x
                          })

# visualize
lymph_viz+
  geom_gate(lymphocyte_gate)

# add lymphocyte gate
gs_pop_add(gs, parent = "live", gate = lymphocyte_gate)
recompute(gs)
```

### Rectangular gate

As the name suggests, we can explicitly define a `rectangular gate` by by defining a `matrix` where the columns are the ranges for markers in `x-axis` and `y-axis`.

```{r, echo = TRUE, eval=TRUE, message=FALSE}
# add T cells gate
## Example of rectangleGate
cd3_vis <- ggcyto(
  gs, subset = "lymphocytes", aes(x = "CD3", y = "CD56")
)+
  geom_hex(bins = 256)+
  facet_wrap(~name)+
  axis_x_inverse_trans()+
  axis_y_inverse_trans()

# visualize
cd3_vis

# using rectangle gate to add T cell gate
cd3_rectanlge <- matrix(c(140, 205, 0, 200),
                       nrow = 2,ncol = 2,
                       byrow = F,
                       dimnames = list(NULL,c("V510-A","U570-A"))) # channel names
cd3_rectangle_gate <- rectangleGate(.gate = cd3_rectanlge,filterId = "CD3+ T cells")
cd3_vis+geom_gate(cd3_rectangle_gate)

# add gate
gs_pop_add(gs, gate = cd3_rectangle_gate, parent = "lymphocytes")
recompute(gs)
```

### Polygon gate

Similar to `Rectangular gate` we can define a `Polygon gate` by generating a `matrix` of **vertices**. Each row of the matrix is the coordinate for a vertex and the columns are the axes.

```{r echo=TRUE,eval=TRUE,message=FALSE}
# add NKT cell gate
## Example of polygonGate
nkt_vis <- ggcyto(gs, subset = "CD3+ T cells",
                  aes(x = "CD1d", y = "CD3"))+ # fuzzy matching of marker names
  geom_hex(bins = 256)+
  facet_wrap(~name)

# visualize
nkt_vis

# define coordinates
nkt_poly <- matrix(c(
  115,140,
  150,150,
  150,180,
  200,180,
  200,140), # coordinates
  ncol = 2,
  byrow = T,
  dimnames = list(NULL, # rownames 
                  c("R670-A","V510-A")) # colnames
)

# convert to gate
nkt_poly_gate <- polygonGate(nkt_poly,
                             filterId = "NKT cells")
# visualize
nkt_vis + geom_gate(nkt_poly_gate)

# move gate and make smaller
nkt_poly_gate <- flowCore::transform_gate(nkt_poly_gate,
                                    dx = 1,
                                    scale = c(1.05,1.05)
                                    )
nkt_vis + geom_gate(nkt_poly_gate)

# add to gs
gs_pop_add(gs,nkt_poly_gate, parent = "CD3+ T cells")
recompute(gs)
```

### Estimating 1D region using `rangeGate`

Previously we used `gate_quantile` to generate a cut point by providing a specific `quantile`. Another approach to do the same when clear **positive** and **negative** populations exist is to use `rangeGate` from the `flowStats` package.

```{r echo=TRUE,eval=TRUE,message=FALSE}
# Add non-NKT cells
## Example rangeGate

# make a list of sample specific gates
non_nkt <- lapply(gs,
                  function(x){
                    ff <- gh_pop_get_data(x, "CD3+ T cells") # extract data at specific node for cleaner calculation
                    flowStats::rangeGate(ff,
                                stain = "R670-A",
                                filterId = "non-NKT Cells",
                                positive = FALSE,
                                alpha = 0.1
                                )
                  })

# visualize
nkt_vis + geom_gate(non_nkt)
```

It looks like we should edit the gate for sample *4002*.

```{r edit_gate, echo=TRUE, eval=TRUE, message=FALSE}
non_nkt[[3]] <- transform_gate(non_nkt[[3]],dx = 25) # small shift in x
nkt_vis + geom_gate(non_nkt)

# add non_nkt
gs_pop_add(gs, non_nkt,parent = "CD3+ T cells")
recompute(gs)
```

### Estimating a 2D Rectangular gate

Below we show an approach to generate a `Rectangular gate` by using sampled data to estimate cut points and then defining a rectangular region using those cut points.

```{r echo=TRUE,eval=TRUE,message=TRUE}
# identify conventional T cells
t_cell_vis <- ggcyto(
  gs, subset = "non-NKT Cells",
  aes(x = "Vg9", y = "Vd1")
)+geom_hex(bins = 256)+
  facet_wrap(~name)

t_cell_vis

# Example use multiple gating tools and piping
## Estimate gate by sampling the data within the entire GatingSet
conv_t_cell_gate <- gs |> 
                          gs_pop_get_data(y = "non-NKT Cells") |> 
  cytoset_to_flowSet() |>
  (function(x){
    
    set.seed(123)
    sample_n <- 1E3
    all_exprs <- fsApply(x,function(y)exprs(y)[sample(nrow(y),sample_n),])
    all_exprs <- flowFrame(all_exprs)
    attrs = c("min","max")
                               # identify cutpoints
                               g1 <- openCyto::gate_quantile(all_exprs,
                                                  channel = "B515-A",
                                                  prob = 0.98
                                                  )
                               g1_attrs <- sapply(attrs, function(y){
                                 attr(g1,y)
                               })
                               g2 <- openCyto::gate_quantile(all_exprs,
                                                  channel = "G575-A",
                                                  prob = 0.90
                                                  )
                               g2_attrs <- sapply(attrs, function(y){
                                 attr(g2,y)
                               })

                               # get cutpoints
                               g1_cutpoint <- g1_attrs[!is.infinite(g1_attrs)]
                               g2_cutpoint <- g2_attrs[!is.infinite(g2_attrs)]
                               
                               # make rectangleGate
                               rg <- rectangleGate(list("B515-A" = c(g1_cutpoint,-Inf),
                                        "G575-A" = c(g2_cutpoint,-Inf)), 
                                        filterId = "conv_Tcells")
    
  })()
 
# visualize
t_cell_vis+ggcyto:::geom_gate(conv_t_cell_gate)+facet_wrap(~name)

# adjust gates
conv_t_cell_gate <- flowCore::transform_gate(conv_t_cell_gate,
                                             dx = 11, # change x
                                             dy = 11 # change y
                                             )

# visualize
t_cell_vis+ggcyto:::geom_gate(conv_t_cell_gate)+facet_wrap(~name)

# add  the calculated rectangleGate
gs_pop_add(gs, conv_t_cell_gate,parent = "non-NKT Cells")
recompute(gs)
```

### Estimating a Ellipsoid gate for a rare population

Previously we used `gate_flowclust_2d` to estimate a gate which we called `lymphocytes`. Below, we use the same gating tool to estimate a *rare* population of T cells called MAIT Cells in a semi-supervised manner.

```{r echo=TRUE, eval=TRUE,message=FALSE}
# identify MAIT cells
mait_vis <- ggcyto(gs, subset = "conv_Tcells",
                   aes(x  = "TCR Va7_2", y = "CD161"))+
  geom_hex(bins = 256)+
  facet_wrap(~name)

mait_vis

# add MAIT cells
## Example of estimation using collapsed data
mait_gate_all <- openCyto::gate_flowclust_2d(fr =  flowFrame(fsApply(gs_pop_get_data(gs,"conv_Tcells"),
                                                            function(y){
                                                              sample_n = min(nrow(y), 1E4)
                                                              exprs(y)[sample(nrow(y),sample_n),]
                                                              }
                                                            )
                                         ),
                                                  xChannel = "V710-A",
                                                  yChannel = "G660-A",target = c(175,175),
                                                  K = 14,filterId = "MAIT Cells" # K indicates number of clusters to estimate
                                                  
                                         )
# fix filterId
mait_gate_all@filterId <- "MAIT Cells"

# visualize
mait_vis+geom_gate(mait_gate_all)+facet_wrap(~name)

# scale gate
mait_gate_all <- scale_gate(mait_gate_all, scale = 2)

# add gate
gs_pop_add(gs,gate = mait_gate_all, parent = "conv_Tcells")
recompute(gs)
```

### Boolean filtering

We can use `booleanFilter` to simply **negate** a gated population! The resulting filter does not have a geometric representation.

```{r echo=TRUE, eval=TRUE, message=FALSE}

# add not MAIT gate
## Example of booleanFilter
not_mait <- booleanFilter(`!MAIT Cells`, filterId = "not_MAIT")

# add boolean gate
gs_pop_add(gs, not_mait, parent = "conv_Tcells")
recompute(gs)

# visualize
mait_vis+
  geom_overlay(
    gs_pop_get_data(gs, y = "not_MAIT"),
    size = 0.5, 
    colour = "red"
  )
```

*Note:* We are also able to combine multiple gates to generate a `booleanFilter`.

Let's create one for innate like T Cells (NKT cells and MAIT Cells).

```{r}
# using an OR ('|') operator
innate_like_t <- booleanFilter(`NKT cells | MAIT Cells`,
                               filterId = "Innate Like T Cells")
# add population
gs_pop_add(gs, innate_like_t, parent = "CD3+ T cells")
recompute(gs)

# visualize results
ggcyto(gs,
       subset = "Innate Like T Cells",
       aes(x = "CD1d",
            y = "CD161"))+
  geom_hex(bins = 128)
```

### Define a Polygon gate using the results of booleanFilter

Below, we show an approach to generate a `Polygon gate` from the results of `booleanFilter`.

```{r echo=TRUE, eval=TRUE, message=FALSE}
# Example of how to use booleanFilter to create a polygon gate
not_mait_gate <- lapply(seq(1,length(
  gs_pop_get_gate(gs,"MAIT Cells")),1),
  function(x){
    gate = gh_pop_get_gate(gs[[x]],"MAIT Cells")
    # enlarge for stringency
    gate <- scale_gate(gate,3)
    ff <- cytoframe_to_flowFrame(gh_pop_get_data(gs[[x]],"conv_Tcells"))
    fl <- filter(ff,gate)
    idx <- which(!fl@subSet)
    set.seed(123)
    sample_n <- min(length(idx),1E5)
    m.names <- colnames(gate@cov)
    events <- exprs(ff)[idx[sample(length(idx),sample_n)],m.names]
    c.hull <- chull(events)
    polygon_mait <- polygonGate(.gate = events[
      c(c.hull,c.hull[1]),],
      filterId = "not_MAIT_Polygon")
  }
);names(not_mait_gate) <- sampleNames(gs)


# visualize
mait_vis+
  geom_gate(not_mait_gate)

# add gate
gs_pop_add(gs,not_mait_gate,parent = "conv_Tcells")
recompute(gs)
```

### Quad gate

Another useful gate when cells are located (generally) in all 4 quadrants that make up the bi-axial plot. One approach to generate this gate is to use `gate_quad_tmix` from the `openCyto` library.

```{r echo=TRUE, eval=TRUE,message=FALSE}
# visualize CD4 vs CD8
cd4_cd8 <- ggcyto(gs,subset = "not_MAIT",
                  aes(x = "CD4", y = "CD8"))+
  geom_hex(bins = 256)+
  facet_wrap(~name)
cd4_cd8

# add a quad gate
cd4_cd8_quad_gate <- lapply(gs_pop_get_data(gs,"not_MAIT"),
                            function(x){
                              qg <- openCyto::gate_quad_tmix(x, 
                                                             channels = c("U785-A",
                                                                          "V570-A"),
                                                             K = 2)
                              # give readable names
                              r.names <- c("U785-A" = "CD4", # these are human readable names
                                           "V570-A" = "CD8a")
                              for(i in 1:length(qg)){ # iterate over each quadrant
                                 sapply(1:length(r.names),function(x){ # iterate over human readable names
                                    current.name <- attr(qg[[i]],"filterId") # keep a record of the current name
                                    attr(qg[[i]],"filterId") <<- gsub( # substitute and apply to qg
                                      pattern = names(r.names)[x],
                                                           replacement = r.names[x],
                                                           x = current.name)
                                  })
                                  
                              }
                              return(qg) # return modified gate                                     
                            })

# visualize
cd4_cd8+
  geom_gate(cd4_cd8_quad_gate)+
  facet_wrap(~name)

# add gate
gs_pop_add(gs, cd4_cd8_quad_gate,parent = "not_MAIT_Polygon")
recompute(gs)
```

## Visualizing the Gating Hierachy

We have been creating and adding many gates to our `GatingSet`. A useful way to visualize all the gates and their relationship is by plotting the gating Hierarchy as a tree. This gives us an immediate summary of what nodes are present in our `GatingSet`.

```{r echo=TRUE,eval=TRUE,message=FALSE}
# plot the gating tree
plot(gs, bool = TRUE)
```

Another useful approach is to visualize the gated data, with gates that we have generated.

```{r, echo=TRUE, eval=TRUE,message=FALSE}
# visualize the full gating hierarchy
autoplot(gs[[1]],bins = 256, bool = TRUE)+
  ggcyto_par_set(limits = "data") # set data range to be determined by data
```

## Saving our work

Finally, we save our `GatingSet`. The `GatingSet` can be loaded back into R using `load_gs("path/to/a/folder")`.

```{r save_gs, echo=TRUE, eval=FALSE}
# save your work
save_gs(gs, path = "path/to/a/folder")
```

## Conclusion

In this section, we spent some time identifying various types of gates that are available in `cytoverse`. As well, we demonstrated how to programmatically create such gates, either manually (i.e. by defining ranges or vertices) or in a semi-automated and data driven manner.

In [**Part 2**](Gating_2.Rmd) we will take a look at how users can leverage a `gatingTemplate` generate gates as we have done here, but with minimal scripting. It is worth noting that while the use `gatingTemplate` minimizes scripting, it does not compromise on reproducibility!

Lastly, we also made ample use of the `ggcyto` library in order to generate visualizations that helped in QC'ing the data and the gates that we generated. We will go over `ggcyto` in more detail in [**Visualizations**](ggcyto_1-5.Rmd).
