---
title: "Gating Cells in cytoverse: Part 2"
vignette: >
  %\VignetteIndexEntry{Gating Cells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Using openCyto gatingTemplate

In [**Part 1**](Gating_1.Rmd) we identified various gates available in `cytoverse` and how to use them. Here, we show how users can compose a similar gating scheme by using a `gatingTemplate`.

A `gatingTemplate` is a csv file that describes the gating sequence and the gating methods. In the `gatingTemplate` csv file, we can leverage the various automated gating methods that we have previously seen, without explicitly requiring to script it in **R**.

Let's take a look at an example `gatingTemplate`

```{r load_libraries, echo=FALSE, eval=TRUE, message=FALSE,warning=FALSE}
library(flowCore)
library(flowWorkspace)
library(openCyto)
library(CytoverseBioc2023)
cache_workshop_data()

# make gs available
gs <- make_transformed_gs()
```

```{r gatingTemplate_load, echo=TRUE, eval= TRUE, message = FALSE}
# load gatingTemplate csv
gt_file <- get_workshop_data("gating_template/")

knitr::kable(read.csv(gt_file$rpath))
```

### Required columns

There are 10 columns that are required in a gatingTemplate csv file:

-   **alias#** : name (alias) of the gate
-   **pop#** : + or -; a combination such as: --, -+, ++, +-; \*.
-   **parent#** : parent population
-   **dims#** : channels or markernames
-   **gating_method#** : gating method to be used. example: *mindensity*
-   **gating_args**: additional arguments that can be passed to **gating_methods**
-   **collapseDataForGating**: **logical** indicating if the data is to be collapsed onto a single `flowFrame` before estimation
-   **groupBy**: string indicating the grouping variable (within `pData` slot of the `GatingSet`)
-   **preprocessing_method**: preprocessing method to be used
-   **preprocessing_args**: additional arguments that can be passed to **preprocessing_method**

*Note:* \# indicates that these fields must be completed. Others can be left blank.

### Understanding the structure of a `gatingTemplate` csv file

Above, It should be apparent that the each row of the `gatingTemplate` csv file is akin to the function call to a specific type of gate.

Let's take a look at a *simple* example of the node `singlet`:

-   `alias`: singlet *name of the gate*
-   `pop`: + *indicating that we want events inside the gate*
-   `parent`: root *indicating the parent population*
-   `dims`: FSC-A,FSC-H *indicating the channels that the function should use*
-   `gating_method`: singletGate *indicating the gating method*
-   `gating_args`:wider_gate=FALSE,prediction_level=0.95 *indicating a tight gate*
-   `collapseDataForGating`: FALSE *indicating not to collapse the entire `GatingSet` into 1 `flowFrame` for estimation*
-   `groupBy`: NA *indicating that no grouping is to be done*
-   `preprocessing_method`: NA *indicating that no preprocessing is to be done*
-   `preprocessing_args`: NA *indicating that no preprocessing argument is specified*

The row for the `CD4+ T Cells` we note:

-   `gating_method`: refGate

-   `gating_args`: CD4_ref:CD8_ref

A `refGate` (reference gate) allows us to build gates by providing *references* to a previously constructed gates. In this instance, we first constructed 1D gates `CD4_ref` and `CD8_ref`, and leveraged them to generate our `CD4+ T Cells` gate.

Also note:

-   `pop`:+-

-   `dims`: CD4,CD8

Here we indicated that `CD4+ T Cells` are **CD4+** and **CD8-**.

*A word of caution: We did not specify channels, rather we left `openCyto` to match the `dims` argument with markernames of the `GatingSet`. If there are similarly named markers (`CD3` and `CD33`) and we provide `dims: CD3`, the process will not complete successfully!*

## Visualizing the `gatingTemplate` tree

Let us now parse this csv file into a `gatingTemplate` object

```{r gatingTemplate_show, echo=TRUE, eval = TRUE,message=FALSE}
# required library
library(openCyto)

# directly read in a csv
gt <- openCyto::gatingTemplate(gt_file$rpath)
```

Now that we have `gt` we should first visualize the hierarchy to **confirm** that it matches our expectations! To achieve this we use `plot`. As well, we will compare it to the `GatingSet` we created earlier.

```{r plot_gt, echo=TRUE, eval=TRUE, message=FALSE}
# show gt
plot(gt)
```

The tree diagram summarizes the gating path as well as the gating methods used. Also note that the `grey` colored arrows identify **helper gates**. These are gates which constructed in to *help* construct a final gate. Above, the gate for `MAIT Cells` is to be created by first constructing 2, 1D gates `TCR Va7_2+` and `CD161+`.

```{r plot_gs, echo=TRUE, eval=TRUE, message=FALSE}
# show gs
plot(gs)
```

Beside the names, most nodes look the same!

## Adding a new population

Let's append a 1 more population to this scheme: - `Non Naive` defined as `CCR7+CD45RA-` or `CCR7-CD45RA-` or `CCR7-CD45RA+`

To do this, we first create a `Naive` node and then negate it!

```{r append_gt, echo=TRUE, eval=TRUE, message=FALSE}
# Non Naive
additional_rows <- data.frame(
  alias = c("Naive", "Non Naive"),
  pop = c("++","+"),
  parent = c("CD4+ T Cells","CD4+ T Cells"),
  dims = c("CCR7,CD45RA","CCR7,CD45RA"),
  gating_method = c("mindensity", "boolGate"),
  gating_args = c("","!Naive"),
  collapseDataForGating = c("",""),
  groupBy = c("",""),
  preprocessing_method = c("",""),
  preprocessing_args = c("","")
)

gt <- read.csv(gt_file$rpath) |>
  rbind(additional_rows)|> data.table::as.data.table()|>
  gatingTemplate() 

# visualize
plot(gt)
```

## Running and attaching the nodes to the `GatingSet`

Since we have saved our `GatingSet` let's first remove all the nodes including and downstream of `singlet`.

```{r gatingTemplate_run, echo = TRUE, eval = TRUE, message = FALSE, fig.height=10, fig.width=10}
# clean gs
gs_pop_remove(gs, "singlet") # removes singlet and all children nodes
recompute(gs)
```

We now use `gt_gating(gt,gs)` to estimate and attach gates to the `GatingSet` in a single call.

```{r compute_with_gating_template, echo=TRUE, eval=TRUE,message=FALSE,warning=FALSE}
# add gates defined in gatingTemplate
gt_gating(gt, gs) # simply provide the parsed gatingTemplate and the GatingSet
recompute(gs)

# visualize Gating Hierarchy
plot(gs, bool = TRUE)
```

We see that all the helper gates are also visualized. We can't remove them as it would affect downstream nodes! However we can hide them.

```{r hide_nodes_gt, echo=TRUE, eval=TRUE, message=FALSE}
# Too many nodes being shown, let's hide helper gates
gt_toggle_helpergates(gt,gs)

# surface MAIT Cells
## labelled as helper since not_MAIT_Cells was created using reference for MAIT Cells
gs_pop_set_visibility(gs,
                      c("MAIT Cells"),
                      TRUE)

# visualize the full gated data
autoplot(gs[[2]], bool = TRUE, bins = 128)+
  ggcyto_par_set(limits = "data")
```

## Conclusion

In this section we took a brief look at how to use a `gatingTemplate` csv file to compose and rapidly gate cells within a `GatingSet`. The portability of the `gatingTemplate` csv ensures that all users utilizing this template file will result in the exact same `GatingHierarchy` and gates.

There are additional functionalities in the `openCyto` library, which we did not go over due to time constraints. We encourage participants to explore the examples outlined [here](https://cytoverse.org/examples/index.html) as well as read through the [vignette](http://www.bioconductor.org/packages/devel/bioc/vignettes/openCyto/inst/doc/openCytoVignette.html).
