---
title: "Gating Cells in cytoverse"
vignette: >
  %\VignetteIndexEntry{Gating Cells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Background

The purpose of flow cytometry is to make inferences regarding some cell type(s) of interest. Often this involves establishing/drawing a gating heirachy to sequentially filter down to the cell type(s) of interest. This process is very often done manually, and can be very labour intensive. Importantly, manual approach implies large variation when one person does it vs. the next, or even if the same person does it multiple times. The `cytoverse` offers a suite of tools to tackle this problem in a reproducible and programmatic manner.

## GatingSet and GatingHeirarchy

In the previous sections, we have seen (and worked with) `flowFrame`, `flowSet`, `cytoframe`, and `cytoset`. These objects hold the underlying FCS data, allow us to visualize it, manipulate it etc.

In this section we work with `GatingSet` and `GatingHierarchy`. These objects, like the name suggests, store information regarding various gates and filters that we will generate. Importantly, we can save the `GatingSet` which will completely package the analysis as well as the FCS files in an opensource format that can be shared, allowing reproducibility.

First, we will go over variety of native methods that exist in the `cytoverse` which analysts can utilize to gate their cells of interest. Next, we will demonstrate how analysts can automate this process by utilizing a `gatingTemplate` (a csv file that can be used to define and build the hierarchy).

## Building the gating tree

Required libraries

```{r load_library, message=FALSE, echo = TRUE, eval=TRUE}
library(flowWorkspace)
library(ggcyto)
# set ggcyto theme  
theme_set(theme_classic())
library(CytoverseBioc2023)
cache_workshop_data()
```

First we create a `GatingSet`

```{r intro_gatingset, eval = T, echo=TRUE}
# create a cytoset
fcs_files <- get_workshop_data("fcs_data/")
fcs_files_TNK <- dplyr::filter(fcs_files, grepl(pattern = "TNK-CR1",x = rname))
cs <- flowWorkspace::load_cytoset_from_fcs(files = fcs_files_TNK$rpath, # loading a set of FCS files as a cytoset
                                           pattern = "TNK-CR1")

# add mock metadata
set.seed(123)
meta <- data.frame(name = sampleNames(cs), 
                   status = "Healthy", 
                   panel = "T Cell",
                   mock_treatment = sample(x = c("Treated","Control"),size = 4,prob = c(0.5,0.5),replace = TRUE),
                   row.names = sampleNames(cs))
pData(cs) <- meta

# creating a GatingSet
gs <- flowWorkspace::GatingSet(cs)
gs
```

We have now created a `GatingSet` called **gs**. Also note that we have associated some (albeit mock) metadata to the `GatingSet`. We are able to leverage the metadata to help construct gates (if need be). 

First, we revisit some of the concepts that we have seen previously: **compensation** and **transformation**.

We can directly compensate the `GatingSet` object.

```{r comp_trans_gs, echo=TRUE, eval=TRUE, fig.height=10, fig.width=10, message=FALSE,warning=FALSE}
# compensate a GatingSet
spill <- keyword(gs[[1]],"$SPILLOVER") # extract spillover matrix stored within the file

gs <- compensate(gs,spill) # GatingSet will be compensated and stores the spill matrix as well

recompute(gs) # update the gs

# retrieve compensation information; return NULL is GatingSet was not compensated
gs_spill <- gs_get_compensations(gs[[1]])

# output is a rich spillover information
slot(gs_spill[[1]],"spillover")[1:4, 1:4]
```

As well we can transform the `GatingSet`. 
For convenience, we are using a previously defined **transformerList** object. The **transformerList** object was extracted from the workspace file created by the authors of the study[ADD RED]. While we don't go over the process here, please look over [THIS SUPPLEMENTAL VIGNETTE: Small vignette to highlight how one can read a FJ workspace file] for details on how one could do it. 

```{r demo_transformation_gs, echo = TRUE, eval = TRUE}
# transformation of GatingSet
t.rds <- dplyr::filter(get_workshop_data("/fj_wsp"), # a previously defined transformation
                       grepl(pattern = "transform", 
                             x = rname))$rpath
my_trans <- readRDS(file = t.rds)

# make into transformerList object
my_trans_list <- flowWorkspace::transformerList(from = names(my_trans),
                                                trans = my_trans)

gs <- flowWorkspace::transform(gs, my_trans_list) # transforms underlying data
```


## Adding gates and building a hierarchy

Now that we have created a `GatingSet` we should start adding various gates that identifies cells of interest. Each sample within a `GatingSet` is associated with a `GatingHierarchy` that stores information regarding various gates that we would have created and applied to the samples.

Below, we demonstrate various types of gates that can be applied to the samples within a `GatingSet` to build a hierachy.

Note that the gating process is reliant on **visualization** of the data. As such, we will make ample use of the `ggcyto` library, highlighting its usage.

```{r singlet_gate_viz, echo = T, eval = T }
# gates for data cleanup; start by visualizing 
singlet_vis <- autoplot(gs_pop_get_data(gs,"root"), # gs_pop_get_data(gs, node) extracts the underlying data
                        x = "FSC-A",
                        y = "FSC-H",
                        bins = 256
                        )+
              facet_wrap(~sampleNames(gs)) # leverage the metadata that is saved within the GatingSet to facet plots 

singlet_vis
```

Now, we add a gate that is used to filter out **singlets**. For this, we make use of the `flowStats` library from the `cytoverse`.

```{r add_singlets_gate, echo = T, eval = T}
# calculate and add a singlet gate
singlet_gate <- fsApply(gs_pop_get_data(gs),function(x){ # sample wise calculation using fsApply
                flowStats::gate_singlet(x, 
                                        filterId = "singlet", # name for the gate
                                        wider_gate = FALSE, # indicate if the returned gate should be linient
                                        prediction_level = 0.95,
                                        subsample_pct = 0.5 # percent of events to sample to calculate the gate
                                        )
  })

# visualize
singlet_vis + 
  geom_gate(singlet_gate)+ # geom_gate takes a single gate or a list of gates as an argument. list should have the same name as sampleNames(gs)
  facet_wrap(~name)

```

Once visualized, we can add the gates to its respective samples.
*Note:* Singlets gate was estimated for each sample.

```{r add_singlets_gate_2, eval=T, echo=TRUE}
# add gate to the GatingSet 
gs_pop_add(gs, singlet_gate, parent = "root") # add singlet_gate to the root node
recompute(gs) # recompute updates the gs
```

The next few gates are done in a similar manner:

-   visualize marker(s)

-   use `openCyto` or `flowStats` to define gates

-   add to `GatingSet`

Below we add a variety of additional gates to the samples.

```{r additional_gate_examples, echo=TRUE, eval=TRUE}
# calculate a live gate 
## Example of a quantile gate
live_gate <- fsApply(gs_pop_get_data(gs,"singlet"), # indicating that estimation should be done on filtered data: singlet
                     function(x){
                       gate <- openCyto::gate_quantile(fr = x, channel = "U450-A", 
                                                      probs = 0.95,
                                                      filterId = "live"
                                                      )
                       
                       # keep negative events
                       gate@max <- gate@min
                       gate@min <- -Inf
                       return(gate)
                     })

# add live gate
gs_pop_add(gs, parent = "singlet",gate = live_gate)
recompute(gs)

# calculate lymphocyte gate
## Example of using flowClust
lymphocyte_gate <- fsApply(gs_pop_get_data(gs,"live"),
                           function(x){
                             openCyto::gate_flowclust_2d(fr = x,
                                                         xChannel = "FSC-A",
                                                         yChannel = "SSC-A",
                                                         filterId = "lymphocytes",
                                                         K = 1,target = c(1E5,0.5E3))
                           })
# call it lymphocytes
lymphocyte_gate <- lapply(lymphocyte_gate,
                          function(x){
                            x@filterId <- "lymphocytes"
                            x
                          })

# add lymphocyte gate
gs_pop_add(gs, parent = "live", gate = lymphocyte_gate)
recompute(gs)


# add T cells gate
## Example of rectangleGate
cd3_vis <- ggcyto(
  gs, subset = "lymphocytes", aes(x = "CD3", y = "CD56")
)+
  geom_hex(bins = 256)+
  facet_wrap(~name)+
  axis_x_inverse_trans()+
  axis_y_inverse_trans()

# visualize
cd3_vis

# using rectangle gate to add T cell gate
cd3_rectanlge <- matrix(c(140, 205, 0, 200),
                       nrow = 2,ncol = 2,
                       byrow = F,
                       dimnames = list(NULL,c("V510-A","U570-A"))) # channel names
cd3_rectangle_gate <- rectangleGate(.gate = cd3_rectanlge,filterId = "CD3+ T cells")
cd3_vis+geom_gate(cd3_rectangle_gate)

# add gate
gs_pop_add(gs, gate = cd3_rectangle_gate, parent = "lymphocytes")
recompute(gs)

# add NKT cell gate
## Example of polygonGate
nkt_vis <- ggcyto(gs, subset = "CD3+ T cells",
                  aes(x = "CD1d", y = "CD3"))+ # fuzzy matching of marker names
  geom_hex(bins = 256)+
  facet_wrap(~name)

# visualize
nkt_vis

# define coordinates
nkt_poly <- matrix(c(
  115,140,
  150,150,
  150,180,
  200,180,
  200,140), # coordinates
  ncol = 2,
  byrow = T,
  dimnames = list(NULL, # rownames 
                  c("R670-A","V510-A")) # colnames
)

# convert to gate
nkt_poly_gate <- polygonGate(nkt_poly,
                             filterId = "NKT cells")
# visualize
nkt_vis + geom_gate(nkt_poly_gate)

# move gate and make smaller
nkt_poly_gate <- flowCore::transform_gate(nkt_poly_gate,
                                    dx = 1,
                                    scale = c(1.05,1.05)
                                    )
nkt_vis + geom_gate(nkt_poly_gate)

# add to gs
gs_pop_add(gs,nkt_poly_gate, parent = "CD3+ T cells")
recompute(gs)

# Add non-NKT cells
## Example rangeGate

# make a list of sample specific gates
non_nkt <- lapply(gs,
                  function(x){
                    ff <- gh_pop_get_data(x, "CD3+ T cells") # extract data at specific node for cleaner calculation
                    flowStats::rangeGate(ff,
                                stain = "R670-A",
                                filterId = "non-NKT Cells",
                                positive = FALSE,
                                alpha = 0.1
                                )
                  })

# visualize
nkt_vis + geom_gate(non_nkt)
```


```{r broken-example, echo=TRUE, eval=FALSE}
# sample 4002 looks like the gate wasn't set properly. Edit
non_nkt$`1615fa39c8b_4002_TNK-CR1.fcs` <- transform_gate(non_nkt$`1615fa39c8b_4002_TNK-CR1.fcs`,
                                                         dx = 25) # small shift in x
nkt_vis + geom_gate(non_nkt)

# add non_nkt
gs_pop_add(gs, non_nkt,parent = "CD3+ T cells")
recompute(gs)

# identify conventional T cells
t_cell_vis <- ggcyto(
  gs, subset = "non-NKT Cells",
  aes(x = "Vg9", y = "Vd1")
)+geom_hex(bins = 256)+
  facet_wrap(~name)

# Example use multiple gating tools and piping
## Estimate gate by sampling the data within the entire GatingSet
conv_t_cell_gate <- gs |> 
                          gs_pop_get_data(y = "non-NKT Cells") |> 
  cytoset_to_flowSet() |>
  (function(x){
    
    set.seed(123)
    sample_n <- 1E3
    all_exprs <- fsApply(x,function(y)exprs(y)[sample(nrow(y),sample_n),])
    all_exprs <- flowFrame(all_exprs)
    attrs = c("min","max")
                               # identify cutpoints
                               g1 <- openCyto::gate_quantile(all_exprs,
                                                  channel = "B515-A",
                                                  prob = 0.98
                                                  )
                               g1_attrs <- sapply(attrs, function(y){
                                 attr(g1,y)
                               })
                               g2 <- openCyto::gate_quantile(all_exprs,
                                                  channel = "G575-A",
                                                  prob = 0.90
                                                  )
                               g2_attrs <- sapply(attrs, function(y){
                                 attr(g2,y)
                               })

                               # get cutpoints
                               g1_cutpoint <- g1_attrs[!is.infinite(g1_attrs)]
                               g2_cutpoint <- g2_attrs[!is.infinite(g2_attrs)]
                               
                               # make rectangleGate
                               rg <- rectangleGate(list("B515-A" = c(g1_cutpoint,-Inf),
                                        "G575-A" = c(g2_cutpoint,-Inf)), 
                                        filterId = "conv_Tcells")
    
  })()
 
# visualize
t_cell_vis+ggcyto:::geom_gate(conv_t_cell_gate)+facet_wrap(~name)

# adjust gates
conv_t_cell_gate <- flowCore::transform_gate(conv_t_cell_gate,
                                             dx = 11, # change x
                                             dy = 11 # change y
                                             )

# add  the calculated rectangleGate
gs_pop_add(gs, conv_t_cell_gate,parent = "non-NKT Cells")
recompute(gs)

# identify MAIT cells
mait_vis <- ggcyto(gs, subset = "conv_Tcells",
                   aes(x  = "TCR Va7_2", y = "CD161"))+
  geom_hex(bins = 256)+
  facet_wrap(~name)


# add MAIT cells
## Example of estimation using collapsed data
mait_gate_all <- openCyto::gate_flowclust_2d(fr =  flowFrame(fsApply(gs_pop_get_data(gs,"conv_Tcells"),
                                                            function(y){
                                                              sample_n = min(nrow(y), 1E4)
                                                              exprs(y)[sample(nrow(y),sample_n),]
                                                              }
                                                            )
                                         ),
                                                  xChannel = "V710-A",
                                                  yChannel = "G660-A",target = c(175,175),
                                                  K = 14,filterId = "MAIT Cells" # K indicates number of clusters to estimate
                                                  
                                         )
# fix filterId
mait_gate_all@filterId <- "MAIT Cells"

# visualize
mait_vis+geom_gate(mait_gate_all)+facet_wrap(~name)

# scale gate
mait_gate_all <- scale_gate(mait_gate_all, scale = 2)

# add gate
gs_pop_add(gs,gate = mait_gate_all, parent = "conv_Tcells")
recompute(gs)

# add not MAIT gate
## Example of booleanFilter
not_mait <- booleanFilter(`!MAIT Cells`, filterId = "not_MAIT")

# add boolean gate
gs_pop_add(gs, not_mait, parent = "conv_Tcells")
recompute(gs)

# Example of how to use booleanFilter to create a polygon gate
not_mait_gate <- lapply(seq(1,length(
  gs_pop_get_gate(gs,"MAIT Cells")),1),
  function(x){
    gate = gh_pop_get_gate(gs[[x]],"MAIT Cells")
    # enlarge for stringency
    gate <- scale_gate(gate,3)
    ff <- cytoframe_to_flowFrame(gh_pop_get_data(gs[[x]],"conv_Tcells"))
    fl <- filter(ff,gate)
    idx <- which(!fl@subSet)
    set.seed(123)
    sample_n <- min(length(idx),1E5)
    m.names <- colnames(gate@cov)
    events <- exprs(ff)[idx[sample(length(idx),sample_n)],m.names]
    c.hull <- chull(events)
    polygon_mait <- polygonGate(.gate = events[
      c(c.hull,c.hull[1]),],
      filterId = "not_MAIT_Polygon")
  }
);names(not_mait_gate) <- sampleNames(gs)

# add gate
gs_pop_add(gs,not_mait_gate,parent = "conv_Tcells")
recompute(gs)

# visualize CD4 vs CD8
cd4_cd8 <- ggcyto(gs,subset = "not_MAIT",
                  aes(x = "CD4", y = "CD8"))+
  geom_hex(bins = 256)+
  facet_wrap(~name)
cd4_cd8

# add a quad gate
cd4_cd8_quad_gate <- lapply(gs_pop_get_data(gs,"not_MAIT"),
                            function(x){
                              qg <- openCyto::gate_quad_tmix(x, 
                                                             channels = c("U785-A",
                                                                          "V570-A"),
                                                             K = 2)
                              # give readable names
                              r.names <- c("U785-A" = "CD4", # these are human readable names
                                           "V570-A" = "CD8a")
                              for(i in 1:length(qg)){ # iterate over each quadrant
                                 sapply(1:length(r.names),function(x){ # iterate over human readable names
                                    current.name <- attr(qg[[i]],"filterId") # keep a record of the current name
                                    attr(qg[[i]],"filterId") <<- gsub( # substitute and apply to qg
                                      pattern = names(r.names)[x],
                                                           replacement = r.names[x],
                                                           x = current.name)
                                  })
                                  
                              }
                              return(qg) # return modified gate                                     
                            })

# visualize
cd4_cd8+
  geom_gate(cd4_cd8_quad_gate)+
  facet_wrap(~name)

# add gate
gs_pop_add(gs, cd4_cd8_quad_gate,parent = "not_MAIT_Polygon")
recompute(gs)

# T cell subsets
cd4_t_subset <- ggcyto(gs, subset = "CD4+CD8a-",
                   aes(x = "CD45RA", y = "CCR7"))+
  geom_hex(bins = 256)+
  facet_wrap(~name)

# plot the gating tree
plot(gs, bool = TRUE)

# visualize the full gating hierarchy
final_plot <- autoplot(gs[[1]],bins = 256, bool = TRUE)+
  ggcyto_par_set(limits = "data") # set data range to be determined by data
```

Best to save the `GatingSet` now. 

```{r save_gs, echo=TRUE, eval=FALSE}
# save your work
save_gs(gs, path = "path/to/a/folder")
```


## Using openCyto gatingTemplate 

A gatingTemplate is a csv file which describes the gating methods, gating otf

HOW IS IT CREATED

HERE IS HOW IT IS RUN

NEED TO UPDATE THE CACHE TO INCLUDE A GATING TEMPLATE
```{r gatingTemplate}

```


```{r gatingTemplate_load, echo=TRUE, eval= FALSE, message = FALSE}
# required library
library(openCyto)
# read in gatingTemplate
gt <- openCyto::gatingTemplate("path/to/csv/file")

# visualize the anticipated gating heirarchy
plot(gt)
```

Since we have saved our `GatingSet` let's first remove all the nodes including and downstream of `singlet`.

```{r gatingTemplate_run, echo = TRUE, eval = FALSE, message = FALSE}
# clean gs
gs_pop_remove(gs, "singlet") # removes singlet and all children nodes
recompute(gs)

# add gates defined in gatingTemplate
gt_gating(gt, gs) # simply provide the parsed gatingTemplate and the GatingSet
recompute(gs)

# visualize
plot(gs, bool = TRUE)

autoplot(gs[[1]], bool = TRUE, bins = 128)
```

