---
title: "Visualization (1.5)"
output: rmarkdown::html_vignette
---

We have spent a bit of time going over import, manupulation, compensation, and transformation of flow cytometry data. Importantly, we have used visualizations to highlight the importance of various steps. Here, we dedicate some time to visualization of flow cytometry data using `ggcyto` library from the `cytoverse`

For those who have experience working with `ggplot`, the approach is going to be similar. As such, `ggcyto` provides a easy to use yet powerful approach to visualize flow cytometry data.

It is also important to note that `ggcyto` works with each of `flowFrame`, `flowSet`, `cytoframe`, `cytoset`, `GatingSet`, and `GatingHierarchy` and allows the user to utilize various attributes of each of the object. For instance, **metadata** within `pData` slot in `flowSet`/`cytoset` and `GatingSet` to prepare insightful visualizations, rapidly.

[THE GATINGSET BEING USED AS AN EXAMPLE SEEMS TO HAVE LOST INFORMATION REGARDING TRANSFORMATION. WILL PLOTclass]

```{r required_libraries, echo=T, eval=T}
library(flowCore)
library(ggcyto)
# set theme
theme_set(theme_bw())
library(CytoverseBioc2023)
cache_workshop_data()
```

```{r echo=F, eval=T, message=FALSE}
# will have up generate a gatingset which can be used to highlight inverse transformations
# can I source an R script?
source("../R/workshop_gs.R")
ff <- gh_pop_get_data(gs[[1]],"root",
                      returnType = "flowFrame")
```


## Hit the ground running using `autoplot` 

The `autoplot` method is a convenient for rapid generation of plots with minimal effort.

1. Bivariate plot

```{r autoplot_ff, echo=TRUE, eval=TRUE}
# flowFrame: bivariate plot
autoplot(ff,
         x = "CD4", # markername in human readable format
         y = "R730-A", # markername as colname of flowFrame 
         bins = 256)
```

2. Density plot

```{r autoplot_ff_test, eval=TRUE,echo=TRUE}
# flowFrame: density plot
autoplot(ff,
         x = "CD4")+ # y not results in a density plot
         geom_density(fill = "red") # utilize ggplot geom
```

3. Bivariate plot of the entire `GatingHeirarchy`

```{r gs_autoplot, echo=T, eval=TRUE, message=FALSE, fig.height=10, fig.width=10}
# GatingSet: full GatingHierachy
autoplot(gs[[2]], # single GatingHierachy will plot the entire gating path
        bins = 256)+ # bins control resolution
        ggcyto_par_set(limits = "data")+ # axis limits based on data 
        geom_stats(type = c("gate_name","percent")) # add name of gate and statistic
```

4. Specific node within a `GatingSet`

```{r gs_autoplot_gate, echo=TRUE, eval=TRUE, message = FALSE}
# GatingSet: single gate
autoplot(gs,
         gate = "conv_Tcells", # plot a specific gate for all samples 
         bins = 256)+
         ggcyto_par_set(limits = "data")+
         axis_x_inverse_trans()+ # inverse transform the axes
         axis_y_inverse_trans()
```


*Note*: When working with `GatingSet` or `GatingHierarchy`, the `autoplot` method does not take **x** and **y** parameters. Rather either a **gate* can be specified or if left unspecified, the entire `GatingHierachy` is plotted. 

## Granular control using `ggcyto`

The `ggcyto` method takes us 1 step deeper into granular control when visualizing cytometry data. Using this method, users can now start to make visualizations not afforded by `autoplot`.

1. Bivariate plot for data within a specific node using markers other than those defined for a node

```{r gs_ggcyto, message=FALSE, echo=TRUE,eval=TRUE}
ggcyto(gs,
       subset = "singlet", # extract data at specific node
       aes(x = "CD8a", y = "CD4"))+ # any marker combination 
  geom_hex(bins = 256)+
  facet_wrap(~name)
```

The marker combination "CD8a vs CD4" is used to at the node `gs_get_pop_paths(gs)[8]` to define CD4+ vs CD8+ T cells. Unlike with `autoplot`, `ggcyto` method allows us to look at the data from the point of `subset`. In the above example, `subset = "singlet"`.

2. Adding a gate and stats layer that doesn't exist in the `subset` node

During reporting, it is possible that users would want to see 

```{r ggcyto_gs_add_gate}
# get a gate
cd4_gate <- gs_pop_get_gate(gs,"CD4+CD8a-")

ggcyto(gs,
       subset = "singlet", # extract data at specific node
       aes(x = "CD8a", y = "CD4"))+ # any marker combination 
  geom_hex(bins = 256)+
  facet_wrap(~name)+
  geom_gate(cd4_gate)+ # plot gate
  geom_stats(type = "percent")+ # on the fly calculation of the "gate" event
  axis_x_inverse_trans()+
  axis_y_inverse_trans()
```

*Note*: This population has not been defined in the `GatingSet`! Yet, we are able to connect multiple components to generate an informative report quickly. We now know that CD4+ cells (not sure at this point if they are CD3+ T cells) are ~47% of total `singlet` population that we had defined.

In a similar manner, we can use a newly defined gate to interrogate the `GatingSet`.

```{r ggcyto_gs_add_gate_new, message=FALSE}
new_gate <- flowCore::rectangleGate(
  list("V570-A" = c(150,210),
       "V510-A" = c(140,210)),
  filterId = "CD3+CD8a+ Cells"
)
ggcyto(gs,
       subset = "singlet", # extract data at specific node
       aes(x = "CD3", y = "CD8a"))+ # any marker combination 
  geom_hex(bins = 256)+
  facet_grid(mock_treatment~name)+ # facet according to treatment
  geom_gate(new_gate)+
  ggcyto_par_set(limits = "data")+
  geom_stats(type = "percent")+
  axis_x_inverse_trans()+
  axis_y_inverse_trans()

```

Above we created a gate, did not apply it to a `GatingSet` directly, but are now leveraging the `ggcyto` library to rapidly generate insight regarding CD8a+ CD3+ T cells from the `singlet` node. Moreover, we have leveraged the metadata from `pData` to generate additional insight regarding the data.

### Expression using histograms and densities

A common way to report or view specific population data is to ask whether there is a change in expression levels (fluoresence intensity) across samples or groups. 

Let's take a look at PD1 expression in CD8+ T Cells

```{r ggcyto_histogram}
cd8 <- as.ggplot(ggcyto(gs,
       subset = "CD4-CD8a+",
       aes(x = "PD1"))+
  geom_histogram(bins = 125, fill = "yellow", color = "black")+
  facet_wrap(name~mock_treatment,ncol = 1, scales = "free_y")+
  ggcyto::labs_cyto(labels = "marker")) # labeling only marker
cd8
```

What if you would like to compare 2 different populations side by side?

```{r ggcyto_gs_arrange}
cd4 <- as.ggplot(ggcyto(gs,
              subset = "CD4+CD8a-",
              aes(x = "PD1"))+
  geom_histogram(bins = 125, fill = "blue", colour = "black")+
  facet_wrap(name~mock_treatment,ncol = 1, scales = "free_y")+
  ggcyto::labs_cyto(labels = "marker")) # labeling only marker

gridExtra::grid.arrange(cd4,cd8, ncol =2 )
```

Or, if you would like to compare population on top of each other?

```{r ggcyto_geom_overlay, fig.width=10}
overlay <-ggcyto(gs,
                  subset = "CD4+CD8a-",
                  aes(x = "PD1"))+
  geom_density(fill = "red", alpha = 0.3)+
  geom_overlay(
    data = gs_pop_get_data(gs,"MAIT Cells"), # extracting information regarding a different population
    fill = "black", alpha = 0.5)+
  axis_x_inverse_trans()+
  labs(title = "Comparing expression of PD1 in 2 distinct T cells: CD8a+ (red) and CD161+ Va7_2+ MAIT Cells (black)")+
  facet_wrap(mock_treatment~name, nrow = 1)+
  theme(aspect.ratio = 1)
overlay
```

Seems like MAIT Cells tend to have higher PD1 expression than CD8+ T Cells! This may be an interesting observation that is worth following up on.

## Backgating and overlays

There are instances when you may want to visualize a gated population (children/grandchildren/siblings) together with the parent. This is the idea behind backgating. We have already seen met this approach above using `geom_overlay` where density of 1 node was project on top of density of a different unrelated node.

Below, we go through another example where we ask the question: what is the general expression level of CD4 and CD8a on MAIT cells compared to all CD3+ T cells? 

```{r ggcyto_gs_geom_overlay, fig.width=10}
mait_expression <- ggcyto(gs,
                          subset = "conv_Tcells",
                          aes(x = "CD4", y = "CD8a"))+
                   geom_hex(bins = 256)+
                   geom_overlay(data = 
                                  gs_pop_get_data(gs, "MAIT Cells"),
                                size = 0.5, colour = "red", alpha = 0.5)+
                   axis_x_inverse_trans()+
                   axis_y_inverse_trans()+
                   facet_wrap(~name, nrow = 1)+
  labs(title = "MAIT Cells (red) overlaid on top of all conventional T Cells")+
  theme(aspect.ratio = 1)
mait_expression  
```

Seems like there are a lot of MAIT Cells that are CD8a+ or negative for both CD8a and CD4 compared to CD4+.


# Conclusion

`ggcyto` is an extremely flexible and powerful tool to explore and visualize flow cytometry data. [WRITE MORE!]  
