---
title: "Importing and Basics of working with FCS files"
output: html_notebook
---

Flow cytometry data (regardless of the type of instrument used) is generally saved as a **.fcs** file. This file is populated with various parameters: such as userid, instrumentid, 


## Installing/loading required libraries 

Start by installing and loading the Cytoverse package **flowCore** hosted at **bioconductor.org**. This package allows us to read and manipulate the **.fcs** file in R.

```{r load_libraries, echo=TRUE, eval=TRUE, message=FALSE}
if(!require("BiocManager", quietly = TRUE)){
  install.packages("BioManager")
}
if(!("flowCore" %in% installed.packages()))
  BiocManager::install("flowCore")

library("flowCore")
```
## Reading in FCS files

There are 3 approaches to read in FCS files:

- Read in individual files as **flowFrame**
- Read in a set of FCS files as **flowSet**
- Read in a set of FCS files as **cytoset**

*Note: We will revisit flowSet and cytoset*

The function `read.FCS` is used to read in individual files as a `flowFrame` object. 


```{r read.FCS, eval=TRUE, echo=TRUE, message = FALSE}
ff <- read.FCS("data/fcs_data/4000_TNK-CR1.fcs")

# a flowFrame object
ff
```
The `flowFrame` object has 3 slots where various data is stored. 

- `@exprs` stores the expression matrix (i.e. the collected data). Can be accessed via `exprs(ff)`
- `@parameters` stores information pertaining to channels: channel name, marker description,and data ranges as AnnotatedDataFrame. Can be accessed via `parameters(ff)`
- `@description` stores additional information extracted from the FCS file header. The FCS file header follow ISAC guidelines. Visit [here](https://isac-net.org/page/Data-Standards) for more information. Can be accessed via `description(ff)`

## Working with `flowFrame`

Working with a `flowFrame` object is going to be very similar to working with a `dataframe` in R. 

A few useful definitions that help us get oriented with the underlying data in the `flowFrame` object.

**Channels**: Instrument derived labels of various parameters that were measured. Channels are the column names of the `flowFrame`. Any data generated from the same instrument will have the same (similar) Channel names.
**Markers**: User provided labels for various parameters that were measured. 
For example: Channel name: B710-A, Marker name: CD3. 
Marker names are set by the users, may not be unique across experiments/users. 
**Expression**: Measured values. A matrix where every row is an event (frequently a cell) and every column is a channel. 

Examples: 

```{r channels, echo=TRUE, message=FALSE}
# channels
colnames(ff)

# markernames
markernames(ff)

# expression
exprs(ff)[1:5,1:5]

```

Notice that there is a correspondence between channels, markers, and the expression matrix. i.e. the `names` of the named vector `markernames(ff)` are a subset of the columns of the expression matrix `exprs(ff)` as well as the columns of the `flowFrame`.

```{r correspondance, echo=TRUE,message=FALSE}
# interested marker: CD4
marker_cd4 <- markernames(ff)[grepl(pattern = "CD4 ",x = markernames(ff))]
marker_cd4

# extract expression of cd4
e_cd4 <- exprs(ff)[,names(marker_cd4)]
head(e_cd4)

# subset flowFrame by column
s_ff <- ff[,names(marker_cd4)]
s_ff

# subset flowFrame by row
s2_ff <- ff[1:100,]
s2_ff
```

Notice that the subset operation can be applied directly to the `flowFrame` object so that information regarding the file is preserved.

Often `<-` can be used to manipulate the `flowFrame` object.

Example:
```{r manipulate_flowFrame, echo=TRUE, message=FALSE}
# create a new markername
new_name <- c("U785-A" = "test")

# old markernames
markernames(s_ff)

# set new markername
markernames(s_ff) <- new_name
markernames(s_ff)

# manipulating expression values
# notice the data range 
s_ff
hist(exprs(s_ff))

# transform
exprs(s_ff) <- exprs(s_ff)/100

# notice the data range after transformation
s_ff
hist(exprs(s_ff))

```

Notice that the data range was not updated when we used `<-` to change the underlying expression matrix. A good practice is to use `transform` function to transform the underlying expression matrix. Moreover, `transform` can also be used to add new columns to the `flowFrame`.

Example
```{r transform_flowFrame, echo=TRUE,message=FALSE}
# define the transformation
simple_trans <- function(x)x/100

# create a transformList
# this indicates which columns to transform and what transformation to do
t_list <- flowCore::transformList(c("U785-A"),simple_trans)

# before transformation
range(s_ff)

# after transformation
s_ff_trans <- flowCore::transform(s_ff,
                                  t_list)
range(s_ff_trans)

# add a new column
s_ff_trans <- flowCore::transform(s_ff,"New_col" = simple_trans(`U785-A`))
s_ff_trans

```

## Reading in a set of FCS files as `flowSet` or `cytoset`

In an experimental sense, a single FCS file is not very interesting (of course, this is not always true!). When there are a set of FCS files they can be loaded into R either as a `flowSet` or a `cytoset`. 

**flowSet/cytoset**: A collection of FCS files, preferably, but not necessarily from the same panel/experiment.

*Note: difference between `flowSet` and `cytoset` is the underlying representation of the data. `cytoset` is implemented in the package `flowWorkspace` which we will encounter in more detail later. Additionally, `cytoset` is built upon a collection of `cytoframe`, analogous to `flowFrame`.*

Options to create `flowSet`

```{r flowSet_example, echo=T, message=TRUE}
# steps to create flowSet

# option 1
# load fcs files as a list
fcs_list <- lapply(X = list.files(path = "data/fcs_data/",full.names = TRUE),FUN = read.FCS)

# convert to flowSet
fs <- as(fcs_list,"flowSet")
fs

# option 2 (preferred)
fs <- read.flowSet(path = "data/fcs_data/")
fs
```

Options to create a `cytoset`

```{r cytoset_example, echo=TRUE, message=TRUE}
require(flowWorkspace)
# option 1 (preferred)
cs <- load_cytoset_from_fcs(path = "data/fcs_data/")

# option 2 (convert flowSet to cytoset)
cs <- flowSet_to_cytoset(fs)
cs
```

Generally, each experiment has unique metadata properties that can (and should) be supplied to the `flowSet` and `cytoset`. These can be added to by using `pData(x) <- data.frame`.

```{r show_metadata_fs, echo=TRUE,message=TRUE}
# create metadata
metadata <- data.frame(row.names = sampleNames(fs),
                       Treatment = sample(c("Untreated","Treated"),length(fs), replace = TRUE),
                       panel = ifelse(grepl(pattern = "TNK",x = sampleNames(fs)),"T Cell Panel","Myeloid Panel"))

# prior to providing metadata
pData(fs)

# add metadata 
pData(fs) <- metadata

pData(fs)

```

The benefit having metadata is that we can use many of the sub-setting operations in a metadata specific manner. 
*Note: Many of the sub-setting operation for `flowFrame` is also applicable for `flowSet` and `cytoset`. Important distinction, however, is that subset operation for `cytoset` will not create a copy of the data and will point to the same set on disk.* 

This is much more convenient than going back and forth between the files. 

Example: 

```{r flowset_subset, echo=TRUE, message=TRUE}
# subset by files that have myeloid staining panel
fs_myeloid <- fs[pData(fs)[["panel"]] == "Myeloid Panel"
                 ,]

fs_myeloid
pData(fs_myeloid)
```

Example of `cytoset` sub-setting

```{r cytoset_subset, echo=TRUE,message=TRUE}
# demonstrate how cs point to the same underlying data
range(exprs(cs[[1]])[,"B515-A"])

# subset and show prior to transformation
cs_small <- cs[1]
range(exprs(cs_small[[1]])[,"B515-A"])

# create a transformList
trans <- transformList("B515-A",asinh)
# transform
cs_small <- transform(cs_small,trans)

# after transformation
range(exprs(cs_small[[1]])[,"B515-A"])

# whole cs 
range(exprs(cs[[1]])[,"B515-A"])
```

As you see, the transformation was applied to a subset `cs_small` however the original `cs` was also altered highlighting that both objects were pointing to the same data. To perform a deep copy we can use `realize_view()`

```{r cytoset_deep_copy, echo=TRUE, message=TRUE}
# look at underlying expression
range(exprs(cs[[2]])[,"B515-A"])

# subset and show prior to transformation
cs_small2 <- realize_view(cs[2])
range(exprs(cs_small2[[1]])[,"B515-A"])

# create a transformList
trans <- transformList("B515-A",asinh)
# transform
cs_small2 <- transform(cs_small2,trans)

# after transformation
range(exprs(cs_small2[[1]])[,"B515-A"])

# whole cs 
range(exprs(cs[[2]])[,"B515-A"])
```

Notice that `cs` is left unchanged. 

## Adding additional FCS files to `flowSet` and `cytoset`

Lastly, we can also add additional FCS files to a `flowSet` or `cytoset` using `rbind2(flowSet1,flowFrame)` or `cs_add_cytoframe`.

```{r add_more_fcs, echo=TRUE, eval=T}
# add to flowSet
fs_small <- fs[1]
fs_small

# add a flowFrame
fs_small <- rbind2(fs_small, fs[[2]]) # fs[[2]] results in a flowFrame
fs_small

# add a flowSet
fs_small <- rbind2(fs_small, fs[3]) # fs[3] results in a flowSet
fs_small

# add to cytoset
cs_small <- realize_view(cs[1]) # deep copy
cs_small

# no need to point to cs_small
cs_add_cytoframe(cs = cs_small, sn = "Sample Name", cf = cs[[3]]) # cs[[3]] results in a cytoframe
cs_small
```

